---
title: "Actionable Diagnostics & Clinical Reasoning: Signal Analysis"
subtitle: "Interpreting metagenomic outputs for a neuroleptospirosis case"
author: "Minh-Quan Ton-Ngoc, Hao Chung The"
date: "2026-05-15"
format:
  html:
    toc: true
    toc-depth: 3
    toc-expand: true
    number-sections: true
    code-fold: true
    theme: cosmo
    self-contained: true
    smooth-scroll: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{=html}
<style>
.case-note {
  background: #f8f9fa;
  border-left: 4px solid #0d6efd;
  padding: 1rem 1.2rem;
  border-radius: 0.4rem;
  margin-bottom: 1rem;
}
.manual-note {
  background: #fff8e1;
  border-left: 4px solid #f0ad4e;
  padding: 0.9rem 1rem;
  border-radius: 0.4rem;
  margin: 1rem 0;
}
.analysis-note {
  background: #eef6ff;
  border-left: 4px solid #1f78b4;
  padding: 0.95rem 1.1rem;
  border-radius: 0.4rem;
  margin: 1rem 0;
}
.result-card {
  background: #f6fff8;
  border-left: 4px solid #198754;
  padding: 1rem 1.2rem;
  border-radius: 0.4rem;
  margin: 1rem 0;
}
.debrief-box {
  background: #f7f7fb;
  border: 1px solid #d9dee3;
  border-radius: 0.5rem;
  padding: 1rem 1.2rem;
  margin: 1rem 0;
}
.mcq {
  border: 1px solid #d9dee3;
  border-radius: 0.5rem;
  padding: 1rem;
  margin: 1rem 0 1.25rem 0;
  background: #ffffff;
}
.mcq .question {
  font-weight: 600;
  margin-bottom: 0.75rem;
}
.mcq label {
  display: block;
  margin: 0.35rem 0;
}
.mcq button {
  margin-top: 0.75rem;
  margin-right: 0.5rem;
  padding: 0.35rem 0.75rem;
  border: 1px solid #cfd4da;
  border-radius: 0.35rem;
  background: #f8f9fa;
  color: #212529;
  font-size: 0.95rem;
  line-height: 1.4;
  box-shadow: none;
  cursor: pointer;
}
.mcq button:hover,
.mcq button:focus {
  background: #eef1f4;
  border-color: #adb5bd;
}
.mcq .feedback {
  margin-top: 0.75rem;
  padding: 0.75rem;
  border-radius: 0.35rem;
  display: none;
}
.mcq .feedback.correct {
  display: block;
  background: #e8f5e9;
  color: #1b5e20;
}
.mcq .feedback.incorrect {
  display: block;
  background: #fdecea;
  color: #8a1c1c;
}
.next-step {
  margin-top: 0.75rem;
}
.locked-box {
  border: 2px dashed #adb5bd;
  border-radius: 0.5rem;
  padding: 1rem;
  background: #f8f9fa;
  margin: 1rem 0;
}
.small-text {
  font-size: 0.95rem;
  color: #495057;
}
summary {
  font-weight: 600;
}
</style>

<script>
function checkMCQ(id, correctValue, explanation) {
  const box = document.getElementById(id);
  const selected = box.querySelector('input[type="radio"]:checked');
  const feedback = box.querySelector('.feedback');
  if (!selected) {
    feedback.className = 'feedback incorrect';
    feedback.style.display = 'block';
    feedback.innerHTML = 'Please choose one answer before checking.';
    return;
  }
  if (selected.value === correctValue) {
    feedback.className = 'feedback correct';
    feedback.innerHTML = '<strong>Good choice.</strong> ' + explanation;
  } else {
    feedback.className = 'feedback incorrect';
    feedback.innerHTML = '<strong>Not the best answer.</strong> ' + explanation;
  }
  feedback.style.display = 'block';
}

const mcqFeedback = {
  mcq1: {
    A: "This would not explain the prominent altered mental status, neck stiffness, photophobia, and seizure. Mild transaminitis can occur in systemic infection but is not the dominant syndrome.",
    B: "Correct. Fever plus headache, meningism, altered mental status, and seizure is best summarized as an acute meningoencephalitic syndrome with systemic inflammation.",
    C: "The seizure matters, but it is not isolated. The fever and meningeal signs make an infectious or inflammatory CNS process much more likely than primary epilepsy alone.",
    D: "There is no evidence that liver disease is the primary process. The neurological and meningeal findings should drive the first diagnostic frame."
  },
  mcq2: {
    A: "Negative targeted tests reduce the probability of the organisms tested, but they do not exclude all infections. Culture, PCR panels, and Xpert are incomplete screens.",
    B: "A viral cause remains possible early, but the available PCR panel is negative and the low CSF glucose keeps TB, fungal, unusual bacterial, and inflammatory causes in play.",
    C: "Correct. The first-line workup narrows the field but leaves a serious unexplained CNS infection or inflammation syndrome, which is an appropriate setting for hypothesis-free testing.",
    D: "Autoimmune disease can remain in the differential, but the fever, CSF inflammation, and exposure context mean infection should not be abandoned."
  },
  mcq3: {
    A: "A taxonomy label is not self-interpreting. In clinical metagenomics, the specimen type, strength of signal, background, controls, and clinical syndrome all matter.",
    B: "Correct. The next step is to judge whether the sequencing result is coherent with the patient, the specimen, and the competing diagnoses.",
    C: "The result should change the ranking of the differential. It does not prove everything alone, but it makes some explanations much more plausible than others.",
    D: "Repeating the same low-yield path without updating the diagnostic frame misses the value of metagenomics, which is to narrow uncertainty and guide confirmatory action."
  },
  mcq4: {
    A: "Metagenomics is most useful precisely when the diagnosis is not obvious and routine targeted testing has not resolved the case.",
    B: "Metagenomics supports clinical reasoning; it does not replace specimen choice, differential diagnosis, confirmatory testing, or judgement about false positives.",
    C: "Correct. In difficult CNS infection, metagenomics can help include a plausible pathogen and exclude less coherent paths more quickly, provided the signal is interpreted carefully.",
    D: "Clinical metagenomics is now used for sterile-site and hospital diagnostic problems, although interpretation is more demanding than in many environmental applications."
  }
};

function checkMCQDetailed(id, correctValue) {
  const box = document.getElementById(id);
  const selected = box.querySelector('input[type="radio"]:checked');
  const feedback = box.querySelector('.feedback');
  if (!selected) {
    feedback.className = 'feedback incorrect';
    feedback.style.display = 'block';
    feedback.innerHTML = 'Please choose one answer before checking.';
    return;
  }
  const isCorrect = selected.value === correctValue;
  feedback.className = isCorrect ? 'feedback correct' : 'feedback incorrect';
  feedback.innerHTML = mcqFeedback[id][selected.value];
  feedback.style.display = 'block';
}

function resetMCQ(id) {
  const box = document.getElementById(id);
  box.querySelectorAll('input[type="radio"]').forEach(el => el.checked = false);
  const feedback = box.querySelector('.feedback');
  feedback.style.display = 'none';
  feedback.innerHTML = '';
}
</script>
```

::: analysis-note
<p><strong>Transition to analysis:</strong> The sections below provide the full reproducible debrief behind the simulation. They show how the local files were inspected, why only one specimen survives strict filtering, and why high unclassified signal forces a conservative clinical interpretation.</p>
:::

```{r setup}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(scales)
library(patchwork)
library(phyloseq)
library(vegan)
library(knitr)

data_dir <- file.path("..", "materials", "clinical-reasoning")
readbased_dir <- file.path(data_dir, "Readbased_Analysis")

theme_set(
  theme_bw(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title.position = "plot",
      legend.position = "right"
    )
)

fig_dir <- "figures"
dir.create(fig_dir, showWarnings = FALSE)

save_fig <- function(plot, filename, width = 9, height = 5.5, dpi = 300, save_pdf = TRUE) {
  png_path <- file.path(fig_dir, paste0(filename, ".png"))
  ggsave(png_path, plot = plot, width = width, height = height, dpi = dpi, bg = "white")
  if (isTRUE(save_pdf)) {
    pdf_path <- file.path(fig_dir, paste0(filename, ".pdf"))
    ggsave(pdf_path, plot = plot, width = width, height = height, bg = "white")
  }
}

sample_role_levels <- c(
  "Untreated CSF",
  "DNase-treated CSF",
  "Patient serum",
  "Negative-control serum"
)

sample_role_palette <- c(
  "Untreated CSF" = "#1b9e77",
  "DNase-treated CSF" = "#7570b3",
  "Patient serum" = "#d95f02",
  "Negative-control serum" = "#666666"
)

signal_palette <- c(
  "Unclassified" = "#bdbdbd",
  "Leptospira santarosai" = "#1b9e77",
  "Eukaryotic/host-associated" = "#80b1d3",
  "Other classified" = "#fdb462"
)

rank_cols <- c("domain", "phylum", "class", "order", "family", "genus", "species")

read_csv_char <- function(path) {
  read_csv(path, show_col_types = FALSE, col_types = cols(.default = col_character()))
}

read_tsv_char <- function(path) {
  read_tsv(path, show_col_types = FALSE, col_types = cols(.default = col_character()))
}

extract_last_taxon <- function(x) {
  ifelse(grepl(";", x), sub("^.*;", "", x), x)
}

clean_taxon <- function(x) {
  x <- gsub("^[a-z]__", "", x)
  x[x == ""] <- NA_character_
  x
}

parse_lineage <- function(df) {
  df %>%
    separate(
      lineage,
      into = rank_cols,
      sep = ";",
      fill = "right",
      extra = "drop",
      remove = FALSE
    ) %>%
    mutate(across(all_of(rank_cols), clean_taxon))
}
```

# Clinical context and adaptation

The original NEJM paper described a 14-year-old boy with severe combined immunodeficiency who had a prolonged, difficult-to-diagnose meningoencephalitic illness. Conventional testing remained unrevealing, and unbiased sequencing of CSF eventually identified *Leptospira*, leading to targeted therapy and clinical recovery.

For this workshop, that story has been adapted into a **Vietnam-facing hospital teaching case**. The educational aim is not only to say "the pathogen was found," but to help healthcare workers reason through a more realistic diagnostic pathway:

-   broad CNS infection differential at admission,
-   staged release of clinical clues,
-   sequencing as a tool for narrowing uncertainty,
-   careful interpretation of which samples are and are not trustworthy,
-   and final integration of sequencing with clinical management.

This notebook now implements that staged pedagogy directly as a self-guided exercise, then continues into the reproducible analysis.

# Available input files

```{r paths}
path_inventory <- tibble::tribble(
  ~object, ~path, ~type,
  "Paper summary notes", file.path(data_dir, "nejm_summary.md"), "file",
  "Vietnam-adapted hypotheses", file.path(data_dir, "nejm_vnmese_hypotheses.md"), "file",
  "Original facilitator template used as source context", file.path(data_dir, "instructor_case_metagenomics_leptospira.qmd"), "file",
  "Original learner template used as source context", file.path(data_dir, "student_case_metagenomics_leptospira.qmd"), "file",
  "Sample metadata", file.path(data_dir, "nejm_sampleinformation.csv"), "file",
  "Workshop taxonomy export (with unclassified)", file.path(data_dir, "nejm_with_unclassified.csv"), "file",
  "Workshop taxonomy export (without unclassified)", file.path(data_dir, "nejm_without_unclassified.csv"), "file",
  "Workshop phyloseq export", file.path(data_dir, "nejm.phyloseq"), "dir",
  "Workshop AMR folder", file.path(data_dir, "nejm.amr"), "dir",
  "Previous Day 4 lesson notebook", "vietnamese-healthy-gut-microbiome.qmd", "file",
  "Underlying read-based outputs", readbased_dir, "dir"
) %>%
  mutate(
    exists = if_else(type == "dir", dir.exists(path), file.exists(path))
  )

kable(path_inventory)
```

All required context files were available when this notebook was assembled, and all main analysis inputs are present locally.

# Load metadata and inspect the proposed sample set

```{r metadata}
metadata <- read_csv(file.path(data_dir, "nejm_sampleinformation.csv"), show_col_types = FALSE) %>%
  mutate(
    sample_name = run_accession,
    sample_role = case_when(
      grepl("untreated CSF", experiment_title, ignore.case = TRUE) ~ "Untreated CSF",
      grepl("DNase-treated CSF", experiment_title, ignore.case = TRUE) ~ "DNase-treated CSF",
      grepl("control", experiment_title, ignore.case = TRUE) ~ "Negative-control serum",
      grepl("serum", experiment_title, ignore.case = TRUE) ~ "Patient serum",
      TRUE ~ sample_name
    ),
    sample_role = factor(sample_role, levels = sample_role_levels),
    patient_status = if_else(grepl("control", experiment_title, ignore.case = TRUE), "Control", "Patient"),
    approx_total_reads = if_else(library_layout == "PAIRED", 2L * read_count, read_count),
    clinical_use = case_when(
      sample_role == "Untreated CSF" ~ "Primary diagnostic CNS specimen",
      sample_role == "DNase-treated CSF" ~ "Comparator CNS specimen after DNase treatment",
      sample_role == "Patient serum" ~ "Systemic comparator specimen",
      sample_role == "Negative-control serum" ~ "Negative control for background/context",
      TRUE ~ "Unspecified"
    )
  ) %>%
  arrange(sample_role)

sample_labels <- setNames(
  paste0(as.character(metadata$sample_role), "\n", metadata$sample_name),
  metadata$sample_name
)

metadata %>%
  select(sample_name, sample_role, patient_status, read_count, approx_total_reads, clinical_use, experiment_title)
```

**Initial metadata interpretation**

-   The local re-analysis clearly targets **four runs**:
    -   untreated CSF,
    -   DNase-treated CSF,
    -   patient serum,
    -   and a negative-control serum.
-   This matches the logic of the original paper, where CSF was the key specimen and serum/control provided context.
-   The local `read_count` values are notably smaller than the multi-million read counts reported in the paper, especially for the untreated CSF and control serum. Whether these reflect SRA spots, processed subsets, or downsampled inputs, they are an important clue that the recreated workflow is **not operating on exactly the same data volume** as the paper.

# Part A: quick interpretation from direct final results

This first interpretation layer uses the direct final results from:

``` text
../materials/clinical-reasoning/Readbased_Analysis/Sourmash-YACHT/final_results/final_results
```

This is the easier, more obvious layer of interpretation. Before running the code, predict what you expect to see:

-   Will all four samples produce a confident pathogen-level result?
-   Which specimen should be most diagnostically useful for CNS infection?
-   Would serum be expected to show the same pathogen signal?

::: {.callout-tip collapse="true"}
## Click to hint

For a CNS infection, CSF is usually more informative than serum. Also remember that a negative control is not expected to produce a disease-causing pathogen signal.
:::

## Inspect the direct final result files

```{r direct-final-files}
direct_final_dir <- file.path(
  readbased_dir, "Sourmash-YACHT",
  "final_results", "final_results"
)

direct_final_files <- tibble(
  file = list.files(direct_final_dir, full.names = FALSE),
  path = file.path(direct_final_dir, file),
  size_bytes = file.info(path)$size
)

kable(direct_final_files)
```

This code follows the same practical style as the previous Vietnamese healthy gut microbiome notebook: first check what files exist, then load the smallest interpretable summary table, then only move to deeper analysis if the first result is incomplete or confusing.

```{r direct-final-results}
direct_with <- read_csv_char(file.path(direct_final_dir, "merged_sourmash_yacht.csv")) %>%
  mutate(
    sample_name = WGS_ID,
    relative_abundance = as.numeric(relative_abundance),
    f_orig_query = as.numeric(f_orig_query),
    total_weighted_hashes = as.numeric(total_weighted_hashes),
    organism_name = coalesce(organism_name, "unclassified")
  ) %>%
  left_join(metadata %>% select(sample_name, sample_role), by = "sample_name")

direct_remove_unclassified <- read_csv_char(file.path(direct_final_dir, "merged_sourmash_yacht_removeuncl_mincoverage0.05.csv")) %>%
  mutate(
    sample_name = WGS_ID,
    organism_name = coalesce(organism_name, "No retained organism")
  ) %>%
  left_join(metadata %>% select(sample_name, sample_role), by = "sample_name")

kable(
  direct_with %>%
    select(sample_name, sample_role, organism_name, lineage, relative_abundance, f_orig_query, total_weighted_hashes, file_name),
  digits = 4
)
```

### Learner checkpoint

Before opening the interpretation, answer:

1.  How many samples show a meaningful retained organism-level result?
2.  Which organism is detected?
3.  Does this match the pathogen described in the original paper?

::: {.callout-note collapse="true"}
## Click to reveal interpretation

The direct final result is very simple:

-   Only **one sample** yields a meaningful pathogen-level result.
-   The informative sample is the **untreated CSF** (`SRR1145846`).
-   The retained organism is ***Leptospira santarosai***.
-   This matches the final pathogen interpretation of the original paper.

This is the clearest direct result layer. It gives learners the main diagnostic message quickly, before they explore why the other samples are weak or non-informative.
:::

```{r direct-retained-hit-table}
kable(
  direct_remove_unclassified %>%
    select(sample_name, sample_role, organism_name, file_name),
  digits = 4
)
```

## Brief AMR inspection at the direct-result stage

AMR analysis is available from the processed `rgi-bwt` outputs, but the key question is whether it changes this clinical diagnosis.

```{r direct-amr-brief}
amr_files <- list.files(file.path(data_dir, "nejm.amr"), pattern = "gene_mapping_data[.]txt$", full.names = TRUE)

read_amr_brief <- function(path) {
  read_tsv_char(path) %>%
    mutate(
      sample_name = sub("[.]gene_mapping_data[.]txt$", "", basename(path)),
      `Completely Mapped Reads` = as.numeric(`Completely Mapped Reads`),
      `Average Percent Coverage` = as.numeric(`Average Percent Coverage`)
    )
}

amr_brief <- bind_rows(lapply(amr_files, read_amr_brief)) %>%
  left_join(metadata %>% select(sample_name, sample_role), by = "sample_name")

amr_brief_summary <- amr_brief %>%
  group_by(sample_name, sample_role) %>%
  summarise(
    n_amr_hits = n(),
    total_completely_mapped_reads = sum(`Completely Mapped Reads`, na.rm = TRUE),
    max_average_percent_coverage = max(`Average Percent Coverage`, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(sample_role)

kable(amr_brief_summary, digits = 2)
```

::: {.callout-note collapse="true"}
## Click to reveal AMR interpretation

The AMR output does **not** materially change the diagnosis in this case.

The AMR calls are sparse, low-coverage, and not central to recognizing neuroleptospirosis. They are useful as a demonstration that clinical metagenomic pipelines can return AMR-related outputs, but here they are mainly illustrative rather than clinically decisive.
:::

# Part B: deeper interpretation using processed analysis tables

The quick result above is enough to identify the most likely pathogen. The deeper analysis below asks harder questions:

-   Why does only one sample survive the strict retained-hit layer?
-   Why are unclassified reads so prominent?
-   Why might the recreated workflow look different from the original paper?
-   When should a sample be interpreted as **non-informative** rather than forced into a diagnosis?

Before opening interpretations in this section, try to predict what the table or figure should show.

::: {.callout-tip collapse="true"}
## Click to hint

The important comparison is not only "which taxon is present?" It is also "how much of each sample is unresolved, and does the taxon survive conservative filtering?"
:::

## Inspect the supplied workshop exports

```{r supplied-exports}
strict_with <- read_csv_char(file.path(data_dir, "nejm_with_unclassified.csv")) %>%
  mutate(
    sample_name = as.character(sample_name),
    unweighted_fraction = as.numeric(unweighted_fraction),
    f_weighted_at_rank = as.numeric(f_weighted_at_rank),
    bp_match_at_rank = as.numeric(bp_match_at_rank),
    total_weighted_hashes = as.numeric(total_weighted_hashes),
    source = "supplied_with_unclassified"
  )

strict_without <- read_csv_char(file.path(data_dir, "nejm_without_unclassified.csv")) %>%
  mutate(
    sample_name = as.character(sample_name),
    unweighted_fraction = as.numeric(unweighted_fraction),
    f_weighted_at_rank = as.numeric(f_weighted_at_rank),
    bp_match_at_rank = as.numeric(bp_match_at_rank),
    total_weighted_hashes = as.numeric(total_weighted_hashes),
    source = "supplied_without_unclassified"
  )

phy_sample_data <- read_csv_char(file.path(data_dir, "nejm.phyloseq", "sample_data.csv"))
phy_otu_export <- read_csv_char(file.path(data_dir, "nejm.phyloseq", "otu_table.csv"))
phy_tax_export <- read_csv_char(file.path(data_dir, "nejm.phyloseq", "tax_table.csv"))

export_inventory <- tibble::tribble(
  ~source, ~n_rows, ~n_unique_samples, ~sample_ids,
  "Metadata", nrow(metadata), n_distinct(metadata$sample_name), paste(metadata$sample_name, collapse = ", "),
  "Supplied with_unclassified", nrow(strict_with), n_distinct(strict_with$sample_name), paste(unique(strict_with$sample_name), collapse = ", "),
  "Supplied without_unclassified", nrow(strict_without), n_distinct(strict_without$sample_name), paste(unique(strict_without$sample_name), collapse = ", "),
  "Supplied phyloseq sample_data", nrow(phy_sample_data), n_distinct(phy_sample_data$sample_name), paste(unique(phy_sample_data$sample_name), collapse = ", ")
)

kable(export_inventory)
```

This is the first major data-quality observation:

-   The metadata and AMR folder describe **four runs**.
-   The supplied taxonomy exports and the supplied `phyloseq` directory retain only **one sample: `SRR1145846`**.
-   That means the workshop export in `nejm_analysis/` is a **strict post-filter subset**, not the whole four-sample diagnostic panel.

```{r supplied-export-details}
supplied_phyloseq_status <- tibble(
  file = c("sample_data.csv", "otu_table.csv", "tax_table.csv"),
  rows = c(nrow(phy_sample_data), nrow(phy_otu_export), nrow(phy_tax_export)),
  cols = c(ncol(phy_sample_data), ncol(phy_otu_export), ncol(phy_tax_export))
)

kable(supplied_phyloseq_status)

strict_with %>%
  select(sample_name, lineage, match_name, unweighted_fraction, f_weighted_at_rank, bp_match_at_rank, total_weighted_hashes)
```

**What the supplied strict export tells us**

-   Only the **untreated CSF** sample survives the strict merged export.
-   In that export, the only retained organism-level call is ***Leptospira santarosai***.
-   Even in that sample, the supplied table still shows a large **unclassified component**, which immediately tells us not to over-simplify the interpretation.
-   Because the supplied export collapses the dataset to one sample, it is **not sufficient** on its own for a four-sample diagnostic teaching notebook.

# Reconstruct a four-sample exploratory taxonomy table from the underlying read-based outputs

To preserve the full diagnostic set, we reconstruct a species-level table from the underlying per-sample taxonomic summaries in:

`../materials/clinical-reasoning/Readbased_Analysis/Sourmash-YACHT/fastmultigather/single_sketches/taxmetagenome/`

This is not identical to the supplied strict YACHT export. It is a broader exploratory view that helps explain why only one sample survived the stricter final filter.

```{r raw-taxonomy}
raw_tax_dir <- file.path(
  readbased_dir, "Sourmash-YACHT",
  "fastmultigather", "single_sketches", "taxmetagenome"
)

raw_tax_files <- sort(list.files(
  raw_tax_dir,
  pattern = "_51[.]gather[.]gtdb[+]euks[.]k51[.]sc10k[.]thresh10k[.]summarized[.]csv$",
  full.names = TRUE
))

raw_tax <- bind_rows(lapply(raw_tax_files, read_csv_char)) %>%
  mutate(
    across(c(fraction, f_weighted_at_rank, bp_match_at_rank, query_ani_at_rank, total_weighted_hashes), as.numeric)
  )

raw_species_with <- raw_tax %>%
  filter(rank == "species") %>%
  transmute(
    sample_name = query_name,
    lineage,
    query_md5,
    query_filename,
    match_name = extract_last_taxon(lineage),
    unweighted_fraction = fraction,
    f_weighted_at_rank,
    bp_match_at_rank,
    total_weighted_hashes,
    query_ani_at_rank
  ) %>%
  parse_lineage() %>%
  left_join(metadata %>% select(sample_name, sample_role, patient_status, approx_total_reads, clinical_use), by = "sample_name") %>%
  mutate(
    species_label = case_when(
      lineage == "unclassified" ~ "unclassified",
      !is.na(species) ~ species,
      !is.na(genus) ~ paste(genus, "unresolved species"),
      TRUE ~ extract_last_taxon(lineage)
    ),
    sample_role = factor(sample_role, levels = sample_role_levels)
  )

raw_species_without <- raw_species_with %>%
  filter(lineage != "unclassified") %>%
  group_by(sample_name) %>%
  mutate(
    f_weighted_classified = f_weighted_at_rank / sum(f_weighted_at_rank, na.rm = TRUE),
    unweighted_fraction_classified = unweighted_fraction / sum(unweighted_fraction, na.rm = TRUE)
  ) %>%
  ungroup()

raw_tax_inventory <- tibble::tribble(
  ~source, ~n_rows, ~n_unique_samples, ~sample_ids,
  "Reconstructed exploratory species table", nrow(raw_species_with), n_distinct(raw_species_with$sample_name), paste(sort(unique(raw_species_with$sample_name)), collapse = ", "),
  "Classified-only exploratory species table", nrow(raw_species_without), n_distinct(raw_species_without$sample_name), paste(sort(unique(raw_species_without$sample_name)), collapse = ", ")
)

kable(raw_tax_inventory)
```

The reconstructed exploratory table restores the expected **four-sample** panel. This is the table used below for the whole-case interpretation, while the supplied strict export is still used as evidence that only one sample passed the stricter retained-hit filter.

## Harmonize sample identifiers across metadata, taxonomy, AMR, and strict exports

```{r sample-id-harmonization}
amr_files <- list.files(file.path(data_dir, "nejm.amr"), pattern = "gene_mapping_data[.]txt$", full.names = TRUE)
amr_sample_ids <- sub("[.]gene_mapping_data[.]txt$", "", basename(amr_files))

harmonization_table <- tibble::tribble(
  ~source, ~n_unique_samples, ~sample_ids,
  "Metadata", n_distinct(metadata$sample_name), paste(sort(unique(metadata$sample_name)), collapse = ", "),
  "Supplied strict taxonomy export", n_distinct(strict_with$sample_name), paste(sort(unique(strict_with$sample_name)), collapse = ", "),
  "Reconstructed exploratory species table", n_distinct(raw_species_with$sample_name), paste(sort(unique(raw_species_with$sample_name)), collapse = ", "),
  "Supplied phyloseq export", n_distinct(phy_sample_data$sample_name), paste(sort(unique(phy_sample_data$sample_name)), collapse = ", "),
  "AMR files", length(unique(amr_sample_ids)), paste(sort(unique(amr_sample_ids)), collapse = ", ")
)

kable(harmonization_table)
```

**Identifier harmonization summary**

-   Sample identifiers are consistent across metadata and AMR outputs.
-   The mismatch is not an ID problem. It is a **retention/filtering problem** in the strict taxonomy export.
-   The reconstructed exploratory table is therefore needed to analyze the four-sample diagnostic set end to end.

# Inspect the YACHT threshold behavior

The underlying YACHT outputs help explain why the merged strict export contains only one sample.

```{r yacht-thresholds}
yacht_dir <- file.path(
  readbased_dir, "Sourmash-YACHT",
  "yacht_results", "temp_yacht_results", "csv"
)

read_yacht_file <- function(path) {
  dat <- read_csv_char(path)
  sample_name <- sub("_min_coverage.*$", "", basename(path))
  threshold <- sub("^.*_min_coverage", "", sub("[.]csv$", "", basename(path)))
  if (nrow(dat) == 0) {
    tibble(
      sample_name = sample_name,
      threshold = threshold,
      organism_name = NA_character_,
      in_sample_est = NA,
      p_vals = NA_real_,
      actual_confidence_with_coverage = NA_real_,
      alt_confidence_mut_rate_with_coverage = NA_real_
    )
  } else {
    dat %>%
      transmute(
        sample_name = WGS_ID,
        threshold = threshold,
        organism_name,
        in_sample_est = as.logical(in_sample_est),
        p_vals = as.numeric(p_vals),
        actual_confidence_with_coverage = as.numeric(actual_confidence_with_coverage),
        alt_confidence_mut_rate_with_coverage = as.numeric(alt_confidence_mut_rate_with_coverage)
      )
  }
}

yacht_thresholds <- bind_rows(lapply(
  list.files(yacht_dir, pattern = "min_coverage(0[.]01|0[.]05)[.]csv$", full.names = TRUE),
  read_yacht_file
)) %>%
  mutate(
    threshold = factor(threshold, levels = c("0.01", "0.05")),
    retained_hit = if_else(is.na(organism_name), "No retained hit", organism_name)
  ) %>%
  left_join(metadata %>% select(sample_name, sample_role), by = "sample_name") %>%
  arrange(sample_role, threshold)

kable(
  yacht_thresholds %>%
    select(sample_name, sample_role, threshold, retained_hit, actual_confidence_with_coverage, alt_confidence_mut_rate_with_coverage),
  digits = 3
)
```

**Threshold interpretation**

-   At the looser `min_coverage = 0.01` setting, three non-untreated-CSF samples retain a weak hit to **SAR11 cluster bacterium PRT-SC02**, which is not clinically coherent for this case.
-   At the stricter `min_coverage = 0.05` setting, **only `SRR1145846`** retains a YACHT hit, and that hit is ***Leptospira santarosai***.
-   This is the key reason the workshop export in `nejm_analysis/` collapses to one sample.

::: {.callout-note collapse="true"}
## Click to reveal: why "many aligned reads" can still mean no pathogen call

A large number of nucleotide alignments is not the same thing as a true pathogen signal.

For a clinically persuasive metagenomic result, we need more than "some reads aligned somewhere." We want a signal that is:

-   enriched in the right specimen,
-   concentrated in a biologically plausible organism,
-   absent or much weaker in serum/control samples,
-   robust enough to survive conservative thresholds,
-   and coherent with the patient's syndrome.

In the original paper, the untreated CSF had a dominant *Leptospira* signal among bacterial reads, while serum and control did not. In contrast, many other bacterial assignments were distributed across background-like genera and appeared in comparator or control libraries. Modern stricter workflows can reasonably remove those weak, distributed calls while retaining the untreated-CSF *Leptospira* signal.
:::

::: {.callout-note collapse="true"}
## Click to reveal: why DNase-treated samples can be confusing for bacterial interpretation

DNase treatment was used to enrich for viral sequencing, not to optimize bacterial species calling. It can reduce free host DNA and shift the library composition, so the relative amount of "microbial-looking" material may increase.

However, the original supplementary material also notes that DNase pretreatment can enrich conserved bacterial 16S rRNA material, which makes accurate species-level bacterial identification harder. Therefore, a DNase-treated library can have more bacterial-looking alignments but less reliable pathogen-level specificity.

For this case, the practical interpretation is:

-   untreated CSF = the strongest bacterial pathogen signal,
-   DNase-treated CSF/serum = more prone to non-specific or hard-to-classify signal,
-   serum/control = useful comparators but not diagnostic for the CNS pathogen in this recreated workflow.
:::

# Taxonomic result inspection

## Unclassified signal is the central interpretive issue

Before running the next chunk, predict which samples will be most difficult to interpret.

Ask yourself:

-   Which specimen is most likely to contain low microbial biomass?
-   Which specimen might be dominated by host/background material?
-   Which specimen should be most clinically meaningful if this is a CNS infection?

::: {.callout-tip collapse="true"}
## Click to hint

Do not only look for the organism name. First look at how much of each sample remains unresolved. A sample can contain a taxon label but still be too incomplete or unstable for a reliable conclusion.
:::

```{r unclassified-summary}
unclassified_summary <- raw_species_with %>%
  filter(lineage == "unclassified") %>%
  select(
    sample_name, sample_role, patient_status,
    unweighted_fraction, f_weighted_at_rank,
    bp_match_at_rank, total_weighted_hashes
  ) %>%
  arrange(sample_role)

kable(unclassified_summary, digits = 3)
```

The unclassified burden is the single most important taxonomic warning sign in this recreated dataset:

-   `SRR1145844` DNase-treated CSF: about **98.1% unweighted** and **99.6% weighted** unclassified.
-   `SRR1145845` patient serum: about **98.4% unweighted** and **99.95% weighted** unclassified.
-   `SRR1145846` untreated CSF: about **47.3% unweighted** but **92.5% weighted** unclassified in the broader species summary.
-   `SRR1145847` negative-control serum: about **50.0% unweighted** and **21.0% weighted** unclassified, with the remaining classified portion split across multiple low-level taxa.

This means several samples are **low-information or unstable at organism level**, even before we ask whether any taxon is clinically plausible.

```{r unclassified-figure}
unclassified_plot_data <- unclassified_summary %>%
  mutate(sample_label = factor(sample_labels[sample_name], levels = sample_labels[metadata$sample_name])) %>%
  pivot_longer(
    cols = c(unweighted_fraction, f_weighted_at_rank),
    names_to = "metric",
    values_to = "fraction"
  ) %>%
  mutate(
    metric = recode(
      metric,
      unweighted_fraction = "Unweighted fraction",
      f_weighted_at_rank = "Weighted fraction"
    ),
    percent = 100 * fraction
  )

unclassified_segments <- unclassified_plot_data %>%
  group_by(sample_name, sample_label, sample_role) %>%
  summarise(
    y_min = min(percent),
    y_max = max(percent),
    .groups = "drop"
  )

unclassified_plot <- ggplot(unclassified_plot_data, aes(x = sample_label, y = percent)) +
  geom_segment(
    data = unclassified_segments,
    aes(x = sample_label, xend = sample_label, y = y_min, yend = y_max),
    inherit.aes = FALSE,
    linewidth = 0.8,
    color = "grey55"
  ) +
  geom_point(aes(fill = metric, shape = metric), size = 3.2, color = "black", stroke = 0.35) +
  scale_fill_manual(values = c("Unweighted fraction" = "#4e79a7", "Weighted fraction" = "#e15759")) +
  scale_shape_manual(values = c("Unweighted fraction" = 21, "Weighted fraction" = 24)) +
  scale_y_continuous(labels = label_percent(scale = 1), limits = c(0, 100), expand = expansion(mult = c(0, 0.02))) +
  labs(
    title = "Unclassified signal dominates most recreated samples",
    subtitle = "Points compare the two abundance metrics within each specimen",
    x = NULL,
    y = "Unclassified fraction (%)",
    fill = "Metric",
    shape = "Metric"
  ) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

save_fig(unclassified_plot, "taxonomy_unclassified_fraction", width = 9.5, height = 5.5)
unclassified_plot
```

**Interpretation**

-   High unclassified signal weakens confidence because much of the sample remains unresolved.
-   When a specimen is dominated by unclassified sequence, the classified profile can become **incomplete, unstable, and overly sensitive to small numbers of assigned hashes**.
-   In a low-biomass clinical specimen, this often means the sample is not robust enough for a trustworthy organism-level conclusion, even if a few taxa are nominally present.
-   For this case, the untreated CSF is the only specimen that still retains a clinically coherent pathogen candidate after strict filtering. The others should be treated cautiously or regarded as **non-informative**.

::: {.callout-note collapse="true"}
## Click to reveal: why unclassified can remain high after host removal

Removing the human genome is necessary, but it does not turn the remaining reads into a clean microbial dataset.

After host depletion, the remaining reads may still include:

-   low-quality or low-complexity fragments,
-   short reads that lack taxonomic information,
-   amplification artifacts,
-   reagent or environmental background,
-   host-associated reads that were not removed perfectly,
-   and real biological sequences without close reference genomes in the database.

This is especially common in low-biomass clinical specimens such as CSF. In the original paper, the untreated CSF had millions of raw reads, but only a small fraction were non-human, and an even smaller fraction carried the diagnostic *Leptospira* signal. The clinical question is therefore not "Can we classify everything?" but "Does the classified part contain a coherent, specimen-specific pathogen signal?"
:::

::: {.callout-note collapse="true"}
## Click to reveal: when "no reliable conclusion" is the right answer

A sample may be best interpreted as **non-informative** when:

-   unclassified signal dominates,
-   the classified taxa are scattered across many weak background-like organisms,
-   the signal does not survive stricter thresholds,
-   the same type of signal appears in serum or control samples,
-   or the result does not fit the clinical syndrome or specimen type.

In this recreated workflow, the DNase-treated CSF, patient serum, and negative-control serum do not provide evidence strong enough for a final organism-level conclusion. That is not a failure of learning. It is one of the core lessons of clinical metagenomics.
:::

## Compare strict retained-hit export versus broader exploratory species summaries

```{r strict-vs-exploratory}
strict_comparison <- strict_with %>%
  select(sample_name, lineage, unweighted_fraction, f_weighted_at_rank) %>%
  mutate(view = "Supplied strict merged export") %>%
  bind_rows(
    raw_species_with %>%
      filter(sample_name %in% strict_with$sample_name) %>%
      select(sample_name, lineage, unweighted_fraction, f_weighted_at_rank) %>%
      mutate(view = "Reconstructed raw species summary")
  ) %>%
  arrange(sample_name, view, desc(f_weighted_at_rank))

kable(strict_comparison, digits = 4)
```

For the untreated CSF, both views agree on the key teaching point: ***Leptospira santarosai*** **is the only clinically coherent signal**. However, the exact numeric fractions differ because these files come from **different summary layers** of the pipeline.

That matters for interpretation:

-   The supplied strict export reflects the final, more conservative retained-hit table.
-   The reconstructed species table is broader and noisier, which is useful for teaching why the other samples did not survive stricter filtering.
-   We should therefore compare the two views qualitatively, not pretend that they are numerically interchangeable.

## Diagnostic composition of each sample

```{r diagnostic-composition}
diagnostic_mass <- raw_species_with %>%
  mutate(
    diagnostic_category = case_when(
      lineage == "unclassified" ~ "Unclassified",
      grepl("Leptospira santarosai$", lineage) ~ "Leptospira santarosai",
      grepl("^Eukaryota", lineage) ~ "Eukaryotic/host-associated",
      TRUE ~ "Other classified"
    )
  ) %>%
  group_by(sample_name, sample_role, diagnostic_category) %>%
  summarise(
    unweighted_fraction = sum(unweighted_fraction, na.rm = TRUE),
    f_weighted_at_rank = sum(f_weighted_at_rank, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(unweighted_fraction, f_weighted_at_rank),
    names_to = "metric",
    values_to = "fraction"
  ) %>%
  mutate(
    metric = recode(
      metric,
      unweighted_fraction = "Unweighted fraction",
      f_weighted_at_rank = "Weighted fraction"
    ),
    diagnostic_category = factor(
      diagnostic_category,
      levels = c("Unclassified", "Leptospira santarosai", "Eukaryotic/host-associated", "Other classified")
    ),
    sample_label = factor(sample_labels[sample_name], levels = sample_labels[metadata$sample_name])
  )

diagnostic_mass_plot <- ggplot(diagnostic_mass, aes(x = sample_label, y = fraction, fill = diagnostic_category)) +
  geom_col(width = 0.75, color = "white") +
  facet_wrap(~metric, ncol = 1) +
  scale_fill_manual(values = signal_palette) +
  scale_y_continuous(labels = percent_format(accuracy = 1), expand = expansion(mult = c(0, 0.02))) +
  labs(
    title = "Only the untreated CSF carries a clinically coherent leptospiral signal",
    subtitle = "The other samples are dominated by unclassified signal or diffuse low-level background",
    x = NULL,
    y = "Fraction of sample",
    fill = "Signal category"
  ) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

save_fig(diagnostic_mass_plot, "taxonomy_diagnostic_signal_vs_unclassified", width = 10, height = 7)
diagnostic_mass_plot
```

**Interpretation**

-   The **untreated CSF** is the only sample where leptospiral signal is large enough to remain central to the story.
-   The **DNase-treated CSF** shows only a tiny leptospiral trace in the broader summary and no retained strict YACHT hit.
-   The **patient serum** is overwhelmingly unclassified and does not support a bloodstream leptospiral signal in this recreated workflow.
-   The **negative-control serum** is not cleanly blank in the broader summary, which argues for caution and supports a conservative interpretation of any low-level non-CSF hits.

## Top classified taxa per sample

```{r top-classified-table}
top_classified_table <- raw_species_without %>%
  group_by(sample_name, sample_role) %>%
  slice_max(order_by = f_weighted_classified, n = 5, with_ties = FALSE) %>%
  ungroup() %>%
  select(sample_name, sample_role, species_label, f_weighted_classified, unweighted_fraction, f_weighted_at_rank)

kable(top_classified_table, digits = 4)
```

This table is useful, but it must be read carefully:

-   `f_weighted_classified` describes **composition among the classified component only**.
-   In samples that are nearly all unclassified, a taxon can look "dominant" **within the classified slice** while still being clinically trivial in the original specimen.
-   That is why the top classified taxa in the DNase-treated CSF, serum, and control serum should **not** be treated as actionable pathogen calls.

```{r species-heatmap}
top_species <- raw_species_without %>%
  group_by(species_label) %>%
  summarise(max_classified = max(f_weighted_classified, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(max_classified), species_label) %>%
  slice_head(n = 10) %>%
  pull(species_label)

heatmap_data <- raw_species_without %>%
  filter(species_label %in% top_species) %>%
  select(sample_name, sample_role, species_label, f_weighted_classified) %>%
  complete(
    sample_name = metadata$sample_name,
    species_label = top_species,
    fill = list(f_weighted_classified = 0)
  ) %>%
  left_join(metadata %>% select(sample_name, sample_role), by = "sample_name") %>%
  mutate(
    sample_label = factor(sample_labels[sample_name], levels = sample_labels[metadata$sample_name]),
    species_label = factor(species_label, levels = rev(top_species))
  )

species_heatmap <- ggplot(
  heatmap_data,
  aes(x = sample_label, y = species_label, fill = 100 * f_weighted_classified)
) +
  geom_tile(color = "grey80", linewidth = 0.3) +
  scale_fill_gradientn(
    colours = c("grey98", "#c7e9c0", "#41ab5d", "#005a32"),
    labels = label_number(suffix = "%", accuracy = 1)
  ) +
  labs(
    title = "Among classified signal, untreated CSF is dominated by Leptospira",
    subtitle = "These are classified-only percentages and should not be mistaken for whole-sample dominance",
    x = NULL,
    y = "Species",
    fill = "Classified-only\nabundance"
  ) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

save_fig(species_heatmap, "taxonomy_species_heatmap", width = 10, height = 6)
species_heatmap
```

**Interpretation**

-   This view helps show why the untreated CSF is clinically different from the other specimens.
-   Once we condition on the classified component, ***Leptospira santarosai*** becomes the dominant classified organism in untreated CSF.
-   The other samples do not show a similarly coherent pathogen pattern. Their classified signal is fragmented across taxa that are either implausible, background-like, or too weak to trust.

# Sample-level diagnostic interpretation

```{r sample-level-diagnostic-table}
strict_hits_05 <- yacht_thresholds %>%
  filter(threshold == "0.05") %>%
  transmute(
    sample_name,
    strict_retained_hit = !is.na(organism_name),
    strict_hit_label = if_else(is.na(organism_name), "No retained hit", organism_name),
    strict_confidence = actual_confidence_with_coverage
  )

top_classified <- raw_species_without %>%
  group_by(sample_name) %>%
  slice_max(order_by = f_weighted_classified, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  transmute(
    sample_name,
    top_classified_species = species_label,
    top_classified_f_weighted_classified = f_weighted_classified
  )

sample_diagnostic_summary <- metadata %>%
  select(sample_name, sample_role, patient_status, approx_total_reads, clinical_use) %>%
  left_join(
    raw_species_with %>%
      filter(lineage == "unclassified") %>%
      select(sample_name, unclassified_unweighted = unweighted_fraction, unclassified_weighted = f_weighted_at_rank, total_weighted_hashes),
    by = "sample_name"
  ) %>%
  left_join(strict_hits_05, by = "sample_name") %>%
  left_join(top_classified, by = "sample_name") %>%
  mutate(
    diagnostic_interpretation = case_when(
      sample_role == "Untreated CSF" ~ "Most clinically meaningful sample. Retains a strict Leptospira hit, but still has substantial unresolved signal.",
      unclassified_weighted >= 0.95 ~ "Essentially non-informative at organism level. Dominated by unclassified signal.",
      sample_role == "Negative-control serum" ~ "Background/control sample with unstable mixed low-level signal. Useful mainly as a cautionary comparator.",
      TRUE ~ "No persuasive pathogen-level evidence strong enough for a final conclusion."
    )
  ) %>%
  arrange(sample_role)

kable(sample_diagnostic_summary, digits = 3)
```

## Practical diagnostic conclusions from the four-sample set

### 1. Untreated CSF (`SRR1145846`)

-   This is the **only sample** that retains a confident strict YACHT hit at `min_coverage = 0.05`.
-   That hit is ***Leptospira santarosai***.
-   In the broader species summary, this sample still carries a heavy unclassified fraction, so the result should be treated as clinically meaningful but not artificially perfect.
-   Even so, this is the **most clinically meaningful specimen** in the recreated workflow and the only sample that supports a plausible actionable diagnosis.

### 2. DNase-treated CSF (`SRR1145844`)

-   This sample is overwhelmingly unclassified.
-   Only a tiny low-level leptospiral trace is visible in the broader species summary, and it disappears from the strict retained-hit table.
-   The most appropriate interpretation is **no reliable organism-level conclusion from this sample**.
-   This is also consistent with the teaching point that DNase treatment can be helpful for viral enrichment but may reduce useful bacterial signal.

### 3. Patient serum (`SRR1145845`)

-   This sample is also overwhelmingly unclassified and does not yield a coherent leptospiral result.
-   The low-level classified taxa are not persuasive for a bloodstream explanation.
-   Therefore, the recreated serum does **not** provide trustworthy evidence for a final pathogen-level conclusion.

### 4. Negative-control serum (`SRR1145847`)

-   The strict YACHT export retains no hit, which is reassuring.
-   However, the broader species summary contains a diffuse mixture of low-level taxa rather than a perfectly clean blank.
-   That does **not** support a real alternative pathogen. Instead, it shows why low-biomass/control samples can look unstable and why conservative filtering matters.

# AMR inspection

```{r amr-load}
read_amr <- function(path) {
  dat <- read_tsv_char(path)
  dat %>%
    mutate(
      sample_name = sub("[.]gene_mapping_data[.]txt$", "", basename(path)),
      `Completely Mapped Reads` = as.numeric(`Completely Mapped Reads`),
      `Average Percent Coverage` = as.numeric(`Average Percent Coverage`),
      `Reference Length` = as.numeric(`Reference Length`)
    )
}

amr_data <- bind_rows(lapply(amr_files, read_amr)) %>%
  left_join(metadata %>% select(sample_name, sample_role), by = "sample_name") %>%
  arrange(sample_role, desc(`Completely Mapped Reads`))

amr_summary <- amr_data %>%
  group_by(sample_name, sample_role) %>%
  summarise(
    n_hits = n(),
    total_complete_reads = sum(`Completely Mapped Reads`, na.rm = TRUE),
    max_coverage = max(`Average Percent Coverage`, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(sample_role)

kable(amr_summary, digits = 2)
```

```{r amr-details}
kable(
  amr_data %>%
    select(sample_name, sample_role, `ARO Term`, `Completely Mapped Reads`, `Average Percent Coverage`, `AMR Gene Family`, `Drug Class`),
  digits = 2
)
```

```{r amr-plot}
amr_plot_data <- amr_data %>%
  mutate(
    sample_label = factor(sample_labels[sample_name], levels = sample_labels[metadata$sample_name]),
    aro_label = factor(`ARO Term`, levels = rev(unique(`ARO Term`)))
  )

amr_plot <- ggplot(
  amr_plot_data,
  aes(x = sample_label, y = aro_label)
) +
  geom_point(
    aes(size = `Completely Mapped Reads`, color = `Average Percent Coverage`),
    alpha = 0.9
  ) +
  scale_color_gradientn(colours = c("#d9d9d9", "#fdae6b", "#e6550d", "#a63603")) +
  scale_size_continuous(range = c(2.5, 10)) +
  labs(
    title = "AMR calls are sparse and not central to the diagnostic question",
    x = NULL,
    y = "ARO term",
    size = "Completely\nmapped reads",
    color = "Average %\ncoverage"
  ) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

save_fig(amr_plot, "amr_sparse_signal_summary", width = 10, height = 5.5)
amr_plot
```

**AMR interpretation**

-   The AMR output is **very sparse**.
-   Only one putative AMR feature is associated with the untreated CSF leptospiral sample, and it is supported by only **2 completely mapped reads**.
-   The higher-read AMR signal appears in the DNase-treated CSF, but that same specimen is taxonomically non-informative for the pathogen question.
-   Coverage is low for most calls, and none of these results materially changes the diagnostic reasoning in this case.

The most appropriate conclusion is that the AMR output is **illustrative rather than clinically informative** here.

# Comparison with the original paper

The original paper reported:

-   **475 single-end CSF reads** mapping to *Leptospira* in untreated CSF,
-   **263 single-end reads** in DNase-treated CSF,
-   **0 leptospiral reads in serum**,
-   **0 leptospiral reads in the negative control**,
-   and eventual confirmation of ***Leptospira santarosai*** by targeted follow-up testing.

Before opening the comparison table, propose two reasons why this recreated output might look different from the published report.

::: {.callout-tip collapse="true"}
## Click to hint

Think about what changed between the original study and this workshop re-analysis: preprocessing, host filtering, reference databases, classification algorithms, and thresholds.
:::

```{r paper-comparison}
comparison_table <- tibble::tribble(
  ~Specimen, ~Original_paper, ~Recreated_workflow, ~Teaching_interpretation,
  "Untreated CSF",
  "Clear actionable leptospiral signal; 475 single-end reads and 955 paired-end bacterial reads assigned to Leptospira.",
  "Only specimen with a retained strict YACHT hit. Broad species summary still has heavy unclassified signal, but Leptospira remains the only coherent diagnosis-level candidate.",
  "This is the one specimen that materially narrows the differential.",
  "DNase-treated CSF",
  "Leptospiral reads still present in the paper, though lower than untreated CSF.",
  "No retained strict hit. Exploratory species summary is overwhelmingly unclassified, with only a tiny leptospiral trace.",
  "Best interpreted as non-informative in the recreated workflow.",
  "Patient serum",
  "No leptospiral reads detected.",
  "No retained strict hit. Serum is dominated by unclassified signal and weak background taxa.",
  "Does not add trustworthy organism-level evidence.",
  "Negative-control serum",
  "No leptospiral reads detected and effectively clean for the key pathogen question.",
  "No retained strict hit, but the broader species summary is noisier than expected, with mixed low-level taxa.",
  "Useful as a cautionary control showing why low-level signal should not be over-called."
)

kable(comparison_table)
```

## Main similarities with the paper

-   The **untreated CSF** remains the most important sample.
-   The recreated workflow still points to ***Leptospira santarosai*** in that specimen.
-   Serum does **not** provide stronger competing evidence for an alternative pathogen.
-   The case still teaches the same major clinical lesson: unbiased CSF metagenomics can shorten the path to a treatable diagnosis when serial conventional testing is unrevealing.

## Main differences from the paper

-   The recreated dataset is **much noisier**, especially outside the untreated CSF.
-   The recreated **DNase-treated CSF** is not cleanly supportive, whereas the paper still recovered leptospiral signal there.
-   The recreated **negative-control serum** is not as clean as the paper's description would lead us to expect.
-   The supplied strict workshop export collapses the dataset to **one retained sample**, which is a much more conservative picture than a simplified classroom result card might suggest.

::: {.callout-note collapse="true"}
## Click to reveal: how to interpret disagreement without calling either analysis wrong

This comparison should not be read as "the paper was right and the re-analysis is wrong" or the reverse.

The original paper used the tools and databases available in 2013-2014 and then confirmed the result with PCR, Sanger sequencing, phylogenetics, and clinical response. The recreated workflow uses modern sketch-based and threshold-sensitive tools that may prioritize specificity and may discard weak or distributed background calls.

Both perspectives can be compatible:

-   the clinical conclusion of the original paper remains strongly supported,
-   the modern re-analysis may be stricter about what counts as a retained pathogen-level call,
-   and samples with heavy unclassified/background signal may reasonably become non-informative.
:::

# Why the recreated results may differ so much from the original report

Several factors could plausibly explain the mismatch:

::: {.callout-tip collapse="true"}
## Click to hint before reading the list

Separate biological explanations from workflow explanations. In this case, many differences are likely workflow-related rather than evidence that the pathogen biology was different.
:::

1.  **Different sequencing depth or retained read volume**

The local metadata suggest a much smaller re-analysis input than the original paper described for some specimens, especially untreated CSF and the negative control. Less data means less sensitivity and more instability.

2.  **Different preprocessing and host-removal behavior**

Modern host depletion and stricter preprocessing can reduce false positives, but they can also leave very sparse non-host signal in low-biomass samples.

3.  **Different classification framework**

The original paper used **SURPI** with nt/protein alignment. This notebook uses outputs from **Sourmash/YACHT and taxmetagenome summaries**, which are sketch-based and threshold-sensitive.

4.  **Different databases**

The modern database now contains ***Leptospira santarosai***, so direct species-level matching is easier than it was in 2014. At the same time, different databases can redistribute marginal reads into different taxa.

5.  **Conservative thresholds**

The strict YACHT export uses a `min_coverage = 0.05` retained-hit view. That improves specificity, but it also means borderline samples drop out completely.

6.  **Low-biomass CSF and comparator samples**

Low-biomass clinical metagenomics is especially vulnerable to instability, high host background, stochastic classification, and over-interpretation of tiny signal.

7.  **Unclassified-heavy samples are inherently incomplete**

When most of a sample is unclassified, the organism-level profile is only a partial summary of what remains after filtering and database matching. That is exactly why some recreated samples should be called **non-informative** rather than forced into a diagnosis.

The important teaching message is that disagreement with the original paper does **not** automatically mean the recreated workflow is wrong. It may mean that the recreated workflow is being **more conservative** with weak or unstable samples.

# Integrated clinical interpretation

## Moving from broad differential to sequencing-based narrowing

Before sequencing, the clinically plausible differential remains broad:

-   tuberculous meningitis,
-   viral meningoencephalitis,
-   fungal CNS infection,
-   partially treated bacterial meningitis,
-   autoimmune/inflammatory disease,
-   and less common zoonotic infection.

The sequencing results change that differential unevenly:

-   Most samples **do not** help enough to justify a pathogen-level conclusion.
-   The **untreated CSF** is the exception. It provides the only coherent organism-level signal that survives strict filtering and fits the clinical story.

## Most likely diagnosis

**The most likely final diagnosis is neuroleptospirosis caused by pathogenic *Leptospira*, most consistent with *Leptospira santarosai* in this recreated workflow.**

That conclusion is supported by the combination of:

-   a difficult CNS infection syndrome with repeatedly unrevealing conventional testing,
-   the clinical logic of the original case,
-   the untreated CSF being the best bacterial specimen,
-   a retained strict YACHT hit to ***Leptospira santarosai*** only in untreated CSF,
-   and the lack of a comparably coherent alternative explanation from the other samples.

## Would these recreated results change management?

-   **Untreated CSF:** yes, this result would meaningfully narrow the diagnostic pathway and justify targeted confirmatory discussion.
-   **DNase-treated CSF:** no, this specimen does not provide evidence strong enough for a final organism-level conclusion.
-   **Patient serum:** no, this specimen does not return trustworthy evidence strong enough for a final conclusion.
-   **Negative control:** no, it is informative only as a cautionary background comparator.

# Why some samples may be non-informative

This case is valuable precisely because **not every specimen helps equally**.

In the recreated workflow, a sample should be considered non-informative when:

-   it is dominated by unclassified signal,
-   the classified component is small and fragmented,
-   the top taxa are clinically implausible or inconsistent with the specimen type,
-   the negative control is noisy enough to raise caution about low-level background,
-   or the strict retained-hit step drops the sample completely.

That is exactly what happens for the DNase-treated CSF, patient serum, and control serum here.

In a workshop setting, this is an important lesson: a metagenomic workflow is useful not only when it finds the answer, but also when it shows that a specimen **does not contain sufficiently trustworthy evidence** for a conclusion.

# Limitations

-   This is a **teaching re-analysis**, not a full reproduction of the original study.
-   The supplied strict workshop export retains only **one sample**, so a broader four-sample view had to be reconstructed from lower-level per-sample summaries.
-   The dataset contains only **four runs**, one of which is a negative control.
-   Heavy unclassified signal makes several sample profiles unstable.
-   The abundance metrics differ across summary layers, so exact percentages should not be over-harmonized.
-   The AMR workflow is present but sparse and not clinically central to this case.
-   Diversity and ordination results are purely descriptive and should not be treated as formal microbiome statistics.

# Suggested discussion points for learners

1.  Why is the untreated CSF the most useful sample for this case?
2.  Why does a large unclassified fraction weaken confidence in the DNase-treated CSF and serum results?
3.  Why is a noisy negative control important even when it does not show the final pathogen?
4.  Why can a recreated workflow disagree with a published paper without being obviously wrong?
5.  How should clinicians balance an imperfect but coherent metagenomic result against ongoing diagnostic uncertainty in a deteriorating patient?

# Suggested next steps and follow-up tests

-   Re-review the exposure history for water and animal-associated leptospiral risk.
-   Send targeted **CSF confirmatory molecular testing** for pathogenic *Leptospira* if available.
-   Discuss disease-directed therapy in parallel with confirmatory testing if the full clinical picture is strongly compatible.
-   If clinically appropriate, consider convalescent serology with the clear caveat that immune dysfunction or immunoglobulin replacement can complicate interpretation.
-   Preserve the untreated CSF as the most important specimen for further confirmatory work.

# Optional research-style add-on: ecological summaries

This final add-on is not needed to reach the clinical diagnosis. It is included for learners who want to see how the same processed table can be adapted into a microbiome-style R workflow, similar to the previous Vietnamese healthy gut microbiome notebook.

::: {.callout-warning collapse="false"}
## Important

These plots are descriptive only. With four runs, heavy unclassified signal, and clinical low-biomass specimens, alpha diversity and ordination should **not** be treated as formal microbiome evidence.
:::

## Build an exploratory phyloseq object

The supplied `nejm.phyloseq/` export is too incomplete for four-sample comparisons, so we build a small exploratory object from the reconstructed classified-only species table.

```{r phyloseq-rebuild}
otu_export <- raw_species_without %>%
  select(species_label, sample_name, f_weighted_classified) %>%
  distinct() %>%
  pivot_wider(names_from = sample_name, values_from = f_weighted_classified, values_fill = 0)

tax_export <- raw_species_without %>%
  select(species_label, domain, phylum, class, order, family, genus, species) %>%
  distinct()

sample_export <- metadata %>%
  select(sample_name, sample_role, patient_status, approx_total_reads, clinical_use)

otu_mat <- as.matrix(as.data.frame(otu_export[, -1]))
rownames(otu_mat) <- otu_export$species_label

tax_mat <- as.matrix(as.data.frame(tax_export[, -1]))
rownames(tax_mat) <- tax_export$species_label

sample_df <- as.data.frame(sample_export)
rownames(sample_df) <- sample_df$sample_name
sample_df$sample_name <- NULL

phy_exploratory <- phyloseq(
  otu_table(otu_mat, taxa_are_rows = TRUE),
  tax_table(tax_mat),
  sample_data(sample_df)
)

phy_exploratory
```

## Alpha diversity

Before running this chunk, predict whether the most clinically informative sample should necessarily have the highest diversity.

::: {.callout-tip collapse="true"}
## Click to hint

In low-biomass clinical metagenomics, higher diversity can mean diffuse background, not a better diagnostic specimen.
:::

```{r alpha-diversity}
sample_matrix_alpha <- t(otu_mat)

alpha_diversity <- tibble(
  sample_name = rownames(sample_matrix_alpha),
  Observed = specnumber(sample_matrix_alpha),
  Shannon = diversity(sample_matrix_alpha, index = "shannon")
) %>%
  left_join(metadata %>% select(sample_name, sample_role), by = "sample_name") %>%
  mutate(
    sample_label = factor(sample_labels[sample_name], levels = sample_labels[metadata$sample_name])
  )

kable(alpha_diversity, digits = 3)
```

```{r alpha-plot}
alpha_observed_plot <- ggplot(alpha_diversity, aes(x = sample_label, y = Observed, fill = sample_role)) +
  geom_col(width = 0.75, color = "black") +
  scale_fill_manual(values = sample_role_palette) +
  labs(
    title = "Observed classified species",
    x = NULL,
    y = "Observed species",
    fill = "Sample role"
  ) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

alpha_shannon_plot <- ggplot(alpha_diversity, aes(x = sample_label, y = Shannon, fill = sample_role)) +
  geom_col(width = 0.75, color = "black") +
  scale_fill_manual(values = sample_role_palette) +
  labs(
    title = "Shannon diversity of classified-only signal",
    x = NULL,
    y = "Shannon index",
    fill = "Sample role"
  ) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

alpha_plot <- alpha_observed_plot / alpha_shannon_plot
save_fig(alpha_plot, "taxonomy_alpha_diversity", width = 10, height = 8)
alpha_plot
```

**Interpretation**

-   These alpha diversity values are **descriptive only**.
-   In a clinical low-biomass dataset, higher diversity does **not** automatically mean a more informative specimen. It can just mean "more diffuse background."
-   The untreated CSF is actually **less diverse** than the noisier comparator specimens, which is compatible with a more coherent pathogen-dominated classified component.
-   The DNase-treated CSF, patient serum, and control serum may look more diverse because their classified reads are split across many weak taxa, not because they are clinically richer samples.

## Beta diversity and ordination

```{r ordination}
sample_matrix <- t(otu_mat)
bray_dist <- vegdist(sample_matrix, method = "bray")
ordination <- cmdscale(bray_dist, eig = TRUE, k = 2)

ord_df <- as.data.frame(ordination$points)
colnames(ord_df) <- c("Axis1", "Axis2")
ord_df$sample_name <- rownames(ord_df)

ord_df <- ord_df %>%
  left_join(metadata %>% select(sample_name, sample_role), by = "sample_name") %>%
  mutate(
    sample_label = sample_labels[sample_name]
  )

ord_plot <- ggplot(ord_df, aes(x = Axis1, y = Axis2, color = sample_role, label = sample_label)) +
  geom_hline(yintercept = 0, linewidth = 0.3, color = "grey80") +
  geom_vline(xintercept = 0, linewidth = 0.3, color = "grey80") +
  geom_point(size = 3.5) +
  geom_text(nudge_y = 0.02, size = 3.2, show.legend = FALSE) +
  scale_color_manual(values = sample_role_palette) +
  labs(
    title = "Bray-Curtis ordination of classified-only species profiles",
    subtitle = "Useful as a descriptive picture only; not suitable for formal inference",
    x = "PCoA 1",
    y = "PCoA 2",
    color = "Sample role"
  )

save_fig(ord_plot, "taxonomy_bray_curtis_ordination", width = 9, height = 5.5)
ord_plot
```

**Interpretation**

-   The ordination is best read as a visual summary of how different the classified-only profiles are.
-   It reinforces that the untreated CSF behaves differently from the other samples.
-   However, with only four runs and heavy unclassified signal, this plot is a **teaching visual**, not evidence of a stable ecological pattern.

# Conclusion

This recreated notebook preserves the main clinical-metagenomic lesson of the original NEJM case while being honest about the limits of the local data.

The key conclusion is not that every sample yields a useful answer. It is that **only one sample, the untreated CSF, provides credible organism-level evidence strong enough to narrow the case toward neuroleptospirosis**, while the remaining recreated samples are dominated by unclassified signal or unstable background and should be treated as non-informative.

That is exactly the kind of interpretation healthcare workers need in practice: not just how to spot a plausible pathogen, but also how to recognize when a specimen does **not** support a trustworthy conclusion.

::: {.callout-tip}
Return to the [Day 4 overview](index.qmd).
:::