---
title: "Clinical Foundations & AMR Surveillance: Taxonomy Analysis"
subtitle: "Classified and unclassified signal in a six-sample Vietnamese gut microbiome subset"
author: "Hao Chung The, Minh-Quan Ton-Ngoc"
date: "2026-05-15"
format:
  html:
    toc: true
    toc-depth: 3
    toc-expand: true
    number-sections: true
    code-fold: true
    theme: cosmo
    self-contained: true
    smooth-scroll: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{=html}
<style>
.manual-note {
  background: #fff8e1;
  border-left: 4px solid #f0ad4e;
  padding: 0.9rem 1rem;
  border-radius: 0.4rem;
  margin: 1rem 0;
}
.analysis-note {
  background: #eef6ff;
  border-left: 4px solid #1f78b4;
  padding: 0.95rem 1.1rem;
  border-radius: 0.4rem;
  margin: 1rem 0;
}
.result-card {
  background: #f6fff8;
  border-left: 4px solid #198754;
  padding: 1rem 1.2rem;
  border-radius: 0.4rem;
  margin: 1rem 0;
}
.debrief-box {
  background: #f7f7fb;
  border: 1px solid #d9dee3;
  border-radius: 0.5rem;
  padding: 1rem 1.2rem;
  margin: 1rem 0;
}
.mcq {
  border: 1px solid #d9dee3;
  border-radius: 0.5rem;
  padding: 1rem;
  margin: 1rem 0 1.25rem 0;
  background: #ffffff;
}
.mcq .question {
  font-weight: 600;
  margin-bottom: 0.75rem;
}
.mcq label {
  display: block;
  margin: 0.35rem 0;
}
.mcq button {
  margin-top: 0.75rem;
  margin-right: 0.5rem;
  padding: 0.35rem 0.75rem;
  border: 1px solid #cfd4da;
  border-radius: 0.35rem;
  background: #f8f9fa;
  color: #212529;
  font-size: 0.95rem;
  line-height: 1.4;
  box-shadow: none;
  cursor: pointer;
}
.mcq button:hover,
.mcq button:focus {
  background: #eef1f4;
  border-color: #adb5bd;
}
.mcq .feedback {
  margin-top: 0.75rem;
  padding: 0.75rem;
  border-radius: 0.35rem;
  display: none;
}
.mcq .feedback.correct {
  display: block;
  background: #e8f5e9;
  color: #1b5e20;
}
.mcq .feedback.incorrect {
  display: block;
  background: #fdecea;
  color: #8a1c1c;
}
.next-step {
  margin-top: 0.75rem;
}
.small-text {
  font-size: 0.95rem;
  color: #495057;
}
summary {
  font-weight: 600;
}
code, pre, pre code {
  white-space: pre-wrap;
  overflow-wrap: anywhere;
  word-break: break-word;
}
p code, li code, td code {
  display: inline-block;
  max-width: 100%;
  vertical-align: bottom;
}
.sidebar nav[role="doc-toc"],
.sidebar nav[role="doc-toc"] a,
#TOC,
#TOC a {
  overflow-wrap: anywhere;
  word-break: break-word;
}
</style>

<script>
const mcqFeedbackGut = {
  mcq_scope: {
    A: "That would overstate what six samples can support. The notebook is deliberately framed as a workflow exercise, not a definitive cohort-level replication.",
    B: "Correct. The notebook is meant to teach a compact end-to-end workflow and support cautious directional comparison, not reproduce the entire paper.",
    C: "The workshop absolutely does use real biological data and real outputs. The point is not to avoid biology, but to scale the exercise to something teachable.",
    D: "The paper already exists, so the workshop does not need to repeat every methodological choice exactly. It needs to be explicit about where the teaching workflow differs."
  },
  mcq_setup: {
    A: "A browser is helpful for ENA and HTML reports, but it is not the core requirement that makes the pipeline runnable.",
    B: "Correct. Nextflow needs Java, and the workflow also needs a way to resolve tool environments such as Conda or Apptainer.",
    C: "R is important later for exploratory analysis, but the pipeline itself starts well before the R section.",
    D: "The workshop databases are already prepared. Learners do not need to rebuild them to get started."
  },
  mcq_csv: {
    A: "This is too minimal for the documented samplesheet format. The current wiki expects the short-read columns and a placeholder long-read column.",
    B: "Correct. The documented CSV uses `sample_id,run_id,group,short_reads_1,short_reads_2,long_reads`, with `run_id` optional and `long_reads` left blank here.",
    C: "These column names are close to what some pipelines use, but they do not match the format described in the current metaflow wiki.",
    D: "A TSV can be useful in other contexts, but the documented input here is a CSV samplesheet."
  },
  mcq_config: {
    A: "Those settings matter, but they are already prepared for the workshop and are not the first things learners should edit.",
    B: "Correct. `input` and `outdir` are the two essential path variables learners are expected to adjust before launching the Vietnamese healthy gut microbiome run.",
    C: "Resource settings can be tuned later if needed, but most learners should not begin there.",
    D: "ARG settings are relevant, but the pipeline will not even know what to process if `input` and `outdir` are wrong."
  },
  mcq_ena: {
    A: "The parent study accession is useful context, but it does not itself give you the paired FASTQ URLs you need for sample download.",
    B: "Correct. `fastq_ftp` is the field that exposes the paired download links, while `sample_title` helps you match the correct workshop samples.",
    C: "Coverage and insert size can be interesting later, but they are not the key fields for collecting the raw FASTQ files in this exercise.",
    D: "The publication title helps you orient yourself, but it is not how you recover the read files."
  },
  mcq_outputs: {
    A: "That would hide an important teaching point. The difference between classified-only and with-unclassified views is part of the interpretation lesson.",
    B: "Correct. The two tables answer slightly different questions: one keeps the uncertainty visible, while the other describes composition conditional on classified signal.",
    C: "The issue is not raw speed. The main reason is interpretability and transparency about unclassified signal.",
    D: "The AMR outputs are useful, but they are not the reason for generating both taxonomy tables."
  },
  mcq_unclassified: {
    A: "That would be too strong. A large unclassified fraction means the named taxa summarize only the classified part of the sample, not the whole metagenome.",
    B: "Correct. Once the unclassified fraction is large, the classified-only table is best read as composition conditional on what the database could label.",
    C: "The problem is not solved by removing rows alone. Removing unclassified rows changes the meaning of the remaining relative abundances.",
    D: "High unclassified signal does not invalidate every downstream step, but it does require more careful wording and comparison."
  }
};

function checkMCQGut(id, correctValue) {
  const box = document.getElementById(id);
  const selected = box.querySelector('input[type="radio"]:checked');
  const feedback = box.querySelector('.feedback');
  if (!selected) {
    feedback.className = 'feedback incorrect';
    feedback.style.display = 'block';
    feedback.innerHTML = 'Please choose one answer before checking.';
    return;
  }
  const isCorrect = selected.value === correctValue;
  feedback.className = isCorrect ? 'feedback correct' : 'feedback incorrect';
  feedback.innerHTML = mcqFeedbackGut[id][selected.value];
  feedback.style.display = 'block';
}

function resetMCQGut(id) {
  const box = document.getElementById(id);
  box.querySelectorAll('input[type="radio"]').forEach(el => el.checked = false);
  const feedback = box.querySelector('.feedback');
  feedback.style.display = 'none';
  feedback.innerHTML = '';
}
</script>
```

# R-based exploratory analysis

The sections below assume that the pipeline outputs above already exist. From this point onward, the notebook shifts from <strong>pipeline execution</strong> to <strong>exploratory interpretation</strong>.

```{r}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(ggrepel)
library(phyloseq)
library(vegan)
library(patchwork)
library(scales)
library(data.table)
library(broom)

data_dir <- file.path("..", "materials", "vietnamese-healthy-gut-microbiome")

group_palette <- c(
  "Child 0-23 months" = "#D55E00",
  "Child 24-59 months" = "#0072B2",
  "Adult" = "#009E73"
)

group_shapes <- c(
  "Child 0-23 months" = 16,
  "Child 24-59 months" = 17,
  "Adult" = 15
)

group_labels <- c(
  "Child 0-23 months" = "Child\n(0-23 months)",
  "Child 24-59 months" = "Child\n(24-59 months)",
  "Adult" = "Adult"
)

group_labeller <- labeller(analysis_group = as_labeller(group_labels))

my_color <- c(
  "#4e79a7", "#f28e2b", "#59a14f", "#b6992d",
  "#499894", "#e15759", "#b07aa1", "#9d7660"
)

make_fill_palette <- function(levels_vector, other_label = "Other") {
  core_levels <- setdiff(levels_vector, other_label)
  n_cols <- length(core_levels)
  if (n_cols > length(my_color)) {
    extra_cols <- scales::hue_pal()(n_cols - length(my_color))
    core_cols <- c(my_color, extra_cols)
  } else {
    core_cols <- my_color[seq_len(n_cols)]
  }
  cols <- setNames(core_cols, core_levels)
  if (other_label %in% levels_vector) {
    cols[other_label] <- "grey85"
  }
  cols[levels_vector]
}

paper_amr_class_levels <- c(
  "Aminoglycosides", "β-lactamases", "Fosfomycin", "Fluoroquinolones",
  "Glycopeptides", "MLS", "Chloramphenicol", "Sulfonamides",
  "Tetracyclines", "Trimethoprim", "Other"
)

map_paper_amr_class <- function(drug_class) {
  case_when(
    drug_class %in% c("aminoglycoside antibiotic") ~ "Aminoglycosides",
    drug_class %in% c("cephalosporin", "penicillin beta-lactam", "carbapenem", "monobactam") ~ "β-lactamases",
    drug_class %in% c("phosphonic acid antibiotic") ~ "Fosfomycin",
    drug_class %in% c("fluoroquinolone antibiotic") ~ "Fluoroquinolones",
    drug_class %in% c("glycopeptide antibiotic") ~ "Glycopeptides",
    drug_class %in% c(
      "macrolide antibiotic", "lincosamide antibiotic", "streptogramin antibiotic",
      "streptogramin A antibiotic", "streptogramin B antibiotic"
    ) ~ "MLS",
    drug_class %in% c("phenicol antibiotic") ~ "Chloramphenicol",
    drug_class %in% c("sulfonamide antibiotic") ~ "Sulfonamides",
    drug_class %in% c("tetracycline antibiotic", "glycylcycline") ~ "Tetracyclines",
    drug_class %in% c("diaminopyrimidine antibiotic") ~ "Trimethoprim",
    TRUE ~ "Other"
  )
}

theme_set(
  theme_bw(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title.position = "plot",
      legend.position = "right"
    )
)

fig_dir <- "figures"
dir.create(fig_dir, showWarnings = FALSE)

save_fig <- function(plot, filename, width = 9, height = 5.5, dpi = 300) {
  ggsave(
    filename = file.path(fig_dir, filename),
    plot = plot,
    width = width,
    height = height,
    dpi = dpi,
    bg = "white"
  )
}

parse_lineage <- function(df) {
  df %>%
    tidyr::separate(
      col = lineage,
      into = c("domain", "phylum", "class", "order", "family", "genus", "species"),
      sep = ";",
      fill = "right",
      extra = "drop",
      remove = FALSE
    )
}

trim_sample_ids <- function(x) trimws(as.character(x))
```

**Code note**

-   This setup chunk loads the R packages used throughout the notebook and defines a few helper objects.
-   The notebook now uses **ggplot2 only** for plotting, to keep figure sizing predictable in the rendered output.
-   For color, the stacked bar plots use a richer custom palette, while group-level plots still combine color and point shape so they remain readable on screen and when printed.
-   The default ggplot2 theme is `theme_bw()`, so any plot built directly with ggplot2 follows the workshop style guide.

# Context from the original paper

Before looking at the 6 local samples, we extract the main analytical context from the published article and the local summary notes.

**Paper-level findings that are reasonable to compare against in a workshop subset**

-   The study analyzed **42 healthy Vietnamese participants** split into **0-23 months**, **2-5 years**, and **adults**.
-   The paper reported a **clear age-related microbiome structure difference**, with children under 2 years showing a distinct gut community structure.
-   **Alpha diversity** was lower in the youngest children than in older children and adults.
-   **Beta diversity** showed broad overlap between older children and adults, while younger children were more distinct and more dispersed.
-   The healthy gut resistome was described as a substantial AMR reservoir, with **tetracycline** and **MLS** resistance genes especially common.
-   Genes highlighted in the paper included **ermB**, **lnuC**, **tetM**, **tetO**, **tetQ**, **tetW**, and **tet-32** as near-ubiquitous signals.
-   The paper reported a higher AMR burden in the youngest children, and noted clinically important genes such as **blaCTX-M** and **mphA/mphD** in younger children.

```{r}
paper_context <- tibble::tibble(
  theme = c(
    "Age-related microbiome structure",
    "Alpha diversity",
    "Beta diversity",
    "Broad AMR burden",
    "AMR classes emphasized",
    "AMR genes emphasized"
  ),
  published_finding = c(
    "Children under 2 years had a distinct gut microbiome structure relative to older children and adults.",
    "Youngest children had lower Shannon diversity; older children were closer to adults.",
    "Bray-Curtis ordination showed young children dispersed away from the overlapping older-child and adult cloud.",
    "Youngest children carried the greatest number of detected AMR genes.",
    "Tetracycline and MLS resistance were nearly ubiquitous.",
    "ermB, lnuC, tet-32, tetM, tetO, tetQ, tetW were common; blaCTX-M and mphA/mphD were highlighted as clinically important signals."
  )
)

knitr::kable(paper_context, col.names = c("Theme", "Finding extracted from paper / summary"))
```

# Analysis-ready files used below

The R analysis below assumes that the workflow and post-processing steps have already produced an analysis folder containing:

-   Metadata for the 6 re-analyzed samples.
-   Taxonomy tables with and without unclassified rows.
-   A `vietnamese_gut.phyloseq` folder that contains `otu_table.csv`, `sample_data.csv`, and `tax_table.csv`.
-   A `vietnamese_gut_amr` folder with one `rgi-bwt` gene mapping table per sample.
-   The original paper PDF and a local summary file used only for context.

```{r}
base_dir <- "."

paths <- tibble::tibble(
  object = c(
    "metadata",
    "taxonomy_with_unclassified",
    "taxonomy_without_unclassified",
    "phyloseq_export_dir",
    "amr_dir",
    "local_summary_md"
  ),
  file_name = c(
    "vietnamese_gut_sample_information.csv",
    "vietnamese_gut_with_unclassified.csv",
    "vietnamese_gut_without_unclassified.csv",
    "vietnamese_gut.phyloseq",
    "vietnamese_gut_amr",
    "vietnamese_gut_summary.md"
  )
) %>%
  mutate(
  path = file.path(data_dir, file_name),
  exists = file.exists(path) | dir.exists(path)
)

knitr::kable(paths)
```

# Metadata inspection

## Load data

**Code note**

-   This chunk reads the local sample metadata and creates a simple `analysis_group` variable that we can use consistently in plots and summaries.
-   The grouping is intentionally simple because the metadata for this workshop subset are limited.

```{r}
metadata <- read_csv(file.path(data_dir, "vietnamese_gut_sample_information.csv"), show_col_types = FALSE) %>%
  mutate(
    sample_name = trim_sample_ids(sample_name),
    studycode = trim_sample_ids(studycode),
    status = recode(status, "Children" = "Child", .default = status),
    analysis_group = case_when(
      status == "Adult" ~ "Adult",
      age_month < 24 ~ "Child 0-23 months",
      TRUE ~ "Child 24-59 months"
    ),
    analysis_group = factor(
      analysis_group,
      levels = c("Child 0-23 months", "Child 24-59 months", "Adult")
    )
  ) %>%
  arrange(analysis_group, studycode)

sample_order <- metadata %>%
  arrange(analysis_group, studycode) %>%
  pull(sample_name)

studycode_order <- metadata %>%
  arrange(analysis_group, studycode) %>%
  pull(studycode)

metadata
```

**Metadata notes**

-   `status` is the most reliable top-level grouping variable in the local file.
-   For children, `age_month` is usable as age in months.
-   For adults, the `age_month` column appears to contain values such as 25, 32, and 59, which are much more plausible as adult ages in **years** than months.
-   For that reason, this notebook uses `analysis_group` derived from `status` and child age thresholds, and uses the adult age values only as descriptive metadata.

```{r}
metadata_summary <- metadata %>%
  count(status, analysis_group, name = "n_samples")

knitr::kable(metadata_summary)
```

## Harmonize sample IDs across inputs

**Code note**

-   This chunk checks whether the same sample IDs are present in metadata, taxonomy tables, the exported phyloseq tables, and the AMR result files.
-   Harmonizing IDs early helps avoid silent joins to the wrong sample later in the notebook.
-   We keep `sample_name` for joins, but use `studycode` later for plotting because it is more compact on figure axes.

```{r}
tax_with <- read_csv(file.path(data_dir, "vietnamese_gut_with_unclassified.csv"), show_col_types = FALSE) %>%
  mutate(sample_name = trim_sample_ids(sample_name))

tax_without <- read_csv(file.path(data_dir, "vietnamese_gut_without_unclassified.csv"), show_col_types = FALSE) %>%
  mutate(sample_name = trim_sample_ids(sample_name))

phy_sample_data <- read_csv(file.path(data_dir, "vietnamese_gut.phyloseq", "sample_data.csv"), show_col_types = FALSE) %>%
  mutate(sample_name = trim_sample_ids(sample_name))

amr_files <- list.files(file.path(data_dir, "vietnamese_gut_amr"), pattern = "gene_mapping_data\\.txt$", full.names = TRUE)
amr_file_samples <- sub("\\.gene_mapping_data\\.txt$", "", basename(amr_files))

sample_harmonization <- tibble::tibble(
  source = c("metadata", "taxonomy_with", "taxonomy_without", "phyloseq_export", "amr_files"),
  n_unique_samples = c(
    n_distinct(metadata$sample_name),
    n_distinct(tax_with$sample_name),
    n_distinct(tax_without$sample_name),
    n_distinct(phy_sample_data$sample_name),
    length(unique(amr_file_samples))
  ),
  sample_ids = c(
    paste(sort(unique(metadata$sample_name)), collapse = ", "),
    paste(sort(unique(tax_with$sample_name)), collapse = ", "),
    paste(sort(unique(tax_without$sample_name)), collapse = ", "),
    paste(sort(unique(phy_sample_data$sample_name)), collapse = ", "),
    paste(sort(unique(amr_file_samples)), collapse = ", ")
  )
)

knitr::kable(sample_harmonization)
```

All four sources contain the same six sample IDs, so harmonization is straightforward.

## Inspect the exported phyloseq components

```{r}
phy_otu_export <- read_csv(file.path(data_dir, "vietnamese_gut.phyloseq", "otu_table.csv"), show_col_types = FALSE)
phy_tax_export <- read_csv(file.path(data_dir, "vietnamese_gut.phyloseq", "tax_table.csv"), show_col_types = FALSE)

phyloseq_export_status <- tibble::tibble(
  file = c("sample_data.csv", "otu_table.csv", "tax_table.csv"),
  rows = c(nrow(phy_sample_data), nrow(phy_otu_export), nrow(phy_tax_export)),
  columns = c(ncol(phy_sample_data), ncol(phy_otu_export), ncol(phy_tax_export))
)

knitr::kable(phyloseq_export_status)
```

The `vietnamese_gut.phyloseq` path is not a serialized `.rds` phyloseq object. It is an export directory containing the component tables needed to reconstruct a phyloseq object.

## Inspect the AMR folder layout

```{r}
amr_file_inventory <- tibble::tibble(
  file = basename(amr_files),
  sample_name = sub("\\.gene_mapping_data\\.txt$", "", basename(amr_files)),
  size_bytes = file.info(amr_files)$size
)

knitr::kable(amr_file_inventory)
```

```{r}
amr_example <- fread(amr_files[1], sep = "\t", data.table = FALSE)

knitr::kable(head(amr_example, 8))
```

The AMR files are one table per sample and include:

-   ARO term
-   reference model metadata
-   mapped read counts
-   average coverage
-   AMR gene family
-   drug class
-   resistance mechanism

This is sufficient for descriptive gene-level and drug-class summaries.

# Taxonomy analysis

## Load data

**Code note**

-   Here we load the two taxonomy tables and split the `lineage` string into standard taxonomic ranks.
-   We then join the taxonomy data back to the metadata so every row knows which sample and age group it belongs to.

```{r}
tax_with_parsed <- parse_lineage(tax_with) %>%
  left_join(metadata, by = "sample_name")

tax_without_parsed <- parse_lineage(tax_without) %>%
  left_join(metadata, by = "sample_name")

tax_structure_summary <- tibble::tibble(
  table = c("with_unclassified", "without_unclassified"),
  rows = c(nrow(tax_with_parsed), nrow(tax_without_parsed)),
  unique_samples = c(n_distinct(tax_with_parsed$sample_name), n_distinct(tax_without_parsed$sample_name)),
  unique_matches = c(n_distinct(tax_with_parsed$match_name), n_distinct(tax_without_parsed$match_name))
)

knitr::kable(tax_structure_summary)
```

## Inspect data structure

```{r}
taxonomy_numeric_checks <- bind_rows(
  tax_with_parsed %>%
    group_by(sample_name) %>%
    summarise(
      table = "with_unclassified",
      f_weighted_sum = sum(f_weighted_at_rank, na.rm = TRUE),
      unweighted_sum = sum(unweighted_fraction, na.rm = TRUE),
      .groups = "drop"
    ),
  tax_without_parsed %>%
    group_by(sample_name) %>%
    summarise(
      table = "without_unclassified",
      f_weighted_sum = sum(f_weighted_at_rank, na.rm = TRUE),
      unweighted_sum = sum(unweighted_fraction, na.rm = TRUE),
      .groups = "drop"
    )
) %>%
  arrange(table, sample_name)

knitr::kable(taxonomy_numeric_checks, digits = 4)
```

Two points matter here:

-   `f_weighted_at_rank` sums to approximately 1 within each sample in both taxonomy tables, so it is appropriate for abundance-based ecological summaries.
-   `unweighted_fraction` does **not** behave like a standard compositional percentage and should not be interpreted as if it must sum to 1 or 100.

## Compare with vs without unclassified

**Code note**

-   This section isolates the rows labelled `unclassified` in the full taxonomy table.
-   The goal is to show workshop participants how much of each sample remains taxonomically unresolved before using the renormalized table for ecological summaries.

```{r}
unclassified_summary <- tax_with_parsed %>%
  filter(tolower(lineage) == "unclassified") %>%
  transmute(
    sample_name,
    analysis_group,
    unclassified_f_weighted = f_weighted_at_rank,
    unclassified_unweighted = unweighted_fraction
  ) %>%
  arrange(analysis_group, sample_name)

knitr::kable(unclassified_summary, digits = 4)
```

```{r}
unclassified_plot_data <- unclassified_summary %>%
    pivot_longer(
        cols = c(unclassified_f_weighted, unclassified_unweighted),
        names_to = "metric",
        values_to = "fraction"
    ) %>%
    mutate(
        metric = recode(
            metric,
            unclassified_f_weighted = "f_weighted_at_rank",
            unclassified_unweighted = "unweighted_fraction"
        ),
        fraction_percent = 100 * fraction,
        studycode = factor(metadata$studycode[match(sample_name, metadata$sample_name)], levels = studycode_order)
    )

unclassified_segments <- unclassified_plot_data %>%
    group_by(sample_name, studycode, analysis_group) %>%
    summarise(
        y_min = min(fraction_percent, na.rm = TRUE),
        y_max = max(fraction_percent, na.rm = TRUE),
        .groups = "drop"
    ) %>%
    mutate(studycode = factor(studycode, levels = studycode_order))

unclassified_plot <- ggplot(unclassified_plot_data, aes(x = studycode, y = fraction_percent)) +
    geom_segment(
        data = unclassified_segments,
        aes(x = studycode, xend = studycode, y = y_min, yend = y_max),
        linewidth = 0.7,
        color = "grey55"
    ) +
    geom_point(aes(fill = metric, shape = metric), size = 3.2, color = "black", stroke = 0.35) +
    facet_wrap(~analysis_group, scales = "free_x", labeller = group_labeller) +
    scale_y_continuous(
        labels = label_percent(scale = 1, accuracy = 1),
        limits = c(0, 100),
        expand = expansion(mult = c(0, 0.02))
    ) +
    scale_fill_manual(values = c("f_weighted_at_rank" = "#4e79a7", "unweighted_fraction" = "#e15759")) +
    scale_shape_manual(values = c("f_weighted_at_rank" = 21, "unweighted_fraction" = 24)) +
    labs(
        title = "Large unclassified fractions remain in all six samples",
        # subtitle = "Within each sample, the two points are connected vertically to compare f_weighted_at_rank and unweighted_fraction directly",
        x = "Study code",
        y = "Fraction assigned to 'unclassified'",
        fill = "Metric",
        shape = "Metric"
    ) +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))

save_fig(unclassified_plot, "taxonomy_unclassified_fraction.png", width = 8, height = 5)
unclassified_plot
```

**Take-home note.** The gap between the two points for each sample shows how much the interpretation changes depending on whether we use the weighted classified fraction or the unweighted fraction.

In this subset, the weighted unclassified fraction ranges from roughly **38% to 49%**. That is large enough that analyses using the table without unclassified rows should be interpreted as **composition conditional on classified signal**, not as a complete description of all reads.

::::: {#mcq_unclassified .mcq}
::: question
When a sample has a large unclassified fraction, what is the safest way to read the classified-only composition table?
:::

<label><input type="radio" name="mcq_unclassified" value="A"> A. As a complete picture of the whole metagenome</label> <label><input type="radio" name="mcq_unclassified" value="B"> B. As the composition of the classified portion of the sample, not the entire DNA signal</label> <label><input type="radio" name="mcq_unclassified" value="C"> C. As a table that becomes automatically unbiased once unclassified rows are removed</label> <label><input type="radio" name="mcq_unclassified" value="D"> D. As a result that should always be discarded without further discussion</label> <button onclick="checkMCQGut('mcq_unclassified','B')">Check answer</button> <button onclick="resetMCQGut('mcq_unclassified')">Reset</button>

::: feedback
:::

<details class="next-step">

<summary>Hint</summary>

Ask what information is lost when the database cannot confidently assign a large part of the sample.

</details>
:::::

## Reconstruct a phyloseq object for the demonstration

**Code note**

-   This chunk converts the long-format taxonomy table into the three core pieces needed by `phyloseq`: an OTU table, a taxonomy table, and sample metadata.
-   We do not need phyloseq for every step, but it is useful to show how these data could be reused in a standard microbiome workflow.

```{r}
otu_demo <- tax_without_parsed %>%
  select(match_name, sample_name, f_weighted_at_rank) %>%
  distinct() %>%
  pivot_wider(names_from = sample_name, values_from = f_weighted_at_rank, values_fill = 0)

tax_demo <- tax_without_parsed %>%
  select(match_name, domain, phylum, class, order, family, genus, species) %>%
  distinct()

sample_demo <- metadata %>%
  select(sample_name, studycode, status, age_month, analysis_group)

otu_mat <- as.matrix(as.data.frame(otu_demo[, -1]))
rownames(otu_mat) <- otu_demo$match_name

tax_mat <- as.matrix(as.data.frame(tax_demo[, -1]))
rownames(tax_mat) <- tax_demo$match_name

sample_df <- as.data.frame(sample_demo)
rownames(sample_df) <- sample_df$sample_name
sample_df$sample_name <- NULL

phy_demo <- phyloseq(
  otu_table(otu_mat, taxa_are_rows = TRUE),
  tax_table(tax_mat),
  sample_data(sample_df)
)

tibble::tibble(
  nsamples = nsamples(phy_demo),
  ntaxa = ntaxa(phy_demo),
  rank_names = paste(rank_names(phy_demo), collapse = ", ")
)
```

This reconstructed object is convenient for workshop workflows, but the ecology below still uses the same underlying `f_weighted_at_rank` abundances from `vietnamese_gut_without_unclassified.csv`.

## Composition plots

### Phylum composition

**Code note**

-   We first rank phyla by their total abundance across all six samples.
-   To keep the stacked bar plot readable, we keep the **top 10 phyla** and collapse the remainder into `Other`.
-   The x-axis uses `studycode` rather than `sample_name`, which keeps the figure easier to read.
-   Within each sample, the stack is ordered from **highest abundance at the bottom** to **lower abundance above**, with `Other` forced to the top.
-   The stack order is controlled explicitly so the dominant phyla sit lower in each bar and `Other` remains at the top of each sample.

```{r}
top_phyla <- tax_without_parsed %>%
  group_by(phylum) %>%
  summarise(total_abundance = sum(f_weighted_at_rank, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_abundance)) %>%
  slice_head(n = 10) %>%
  pull(phylum)

phylum_plot_data <- tax_without_parsed %>%
  mutate(phylum_plot = if_else(phylum %in% top_phyla, phylum, "Other")) %>%
  group_by(sample_name, studycode, analysis_group, phylum_plot) %>%
  summarise(rel_abundance = sum(f_weighted_at_rank, na.rm = TRUE), .groups = "drop") %>%
  group_by(sample_name, studycode, analysis_group) %>%
  mutate(stack_order = if_else(phylum_plot == "Other", -1, rank(rel_abundance, ties.method = "first"))) %>%
  ungroup()

phylum_levels <- phylum_plot_data %>%
  group_by(phylum_plot) %>%
  summarise(total_abundance = sum(rel_abundance), .groups = "drop") %>%
  arrange(desc(total_abundance)) %>%
  pull(phylum_plot)

phylum_fill <- make_fill_palette(phylum_levels)

phylum_plot_data <- phylum_plot_data %>%
  mutate(
    studycode = factor(studycode, levels = studycode_order),
    phylum_plot = factor(phylum_plot, levels = phylum_levels)
  )

phylum_plot <- ggplot(phylum_plot_data, aes(x = studycode, y = rel_abundance, fill = phylum_plot, order = stack_order)) +
  geom_col(width = 0.75, color = "grey15", linewidth = 0.15) +
  facet_wrap(~ analysis_group, scales = "free_x", labeller = group_labeller) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, 1),
    expand = expansion(mult = c(0, 0.02))
  ) +
  scale_fill_manual(values = phylum_fill, breaks = phylum_levels) +
  labs(
    title = "Phylum-level composition after removing unclassified rows",
    subtitle = "Top 10 phyla by overall relative abundance; 'Other' is stacked on top within each sample",
    x = "Study code",
    y = "Relative abundance (f_weighted_at_rank)",
    fill = "Phylum"
  ) +
  theme(
    axis.text.x = element_text(angle = 20, hjust = 1),
    legend.position = "right"
  )

save_fig(phylum_plot, "taxonomy_phylum_composition.png", width = 8, height = 5.5)
phylum_plot
```

**Take-home note.** This phylum-level stacked bar plot gives the quickest visual summary of how the dominant classified signal shifts across the six samples.

```{r}
phylum_top10_by_sample <- tax_without_parsed %>%
  group_by(studycode, analysis_group, phylum) %>%
  summarise(rel_abundance = sum(f_weighted_at_rank, na.rm = TRUE), .groups = "drop") %>%
  group_by(studycode, analysis_group) %>%
  arrange(desc(rel_abundance), .by_group = TRUE) %>%
  slice_head(n = 10) %>%
  ungroup()

phylum_top10_by_group <- tax_without_parsed %>%
  group_by(studycode, analysis_group, phylum) %>%
  summarise(rel_abundance = sum(f_weighted_at_rank, na.rm = TRUE), .groups = "drop") %>%
  group_by(analysis_group, phylum) %>%
  summarise(mean_rel_abundance = mean(rel_abundance), .groups = "drop") %>%
  group_by(analysis_group) %>%
  arrange(desc(mean_rel_abundance), .by_group = TRUE) %>%
  slice_head(n = 10) %>%
  ungroup()

knitr::kable(phylum_top10_by_sample, digits = 4, caption = "Top 10 phyla by relative abundance within each sample")
knitr::kable(phylum_top10_by_group, digits = 4, caption = "Top 10 phyla by mean relative abundance within each age group")
```

### Genus composition

**Code note**

-   A stacked bar plot at genus level is harder to read if we include too many genera.
-   We therefore keep the **top 10 genera** across the dataset and collapse the rest into `Other`.
-   This gives a sample-level overview that is more intuitive for workshop participants before moving to the genus heatmap.

```{r}
top_genera_bar <- tax_without_parsed %>%
  filter(!is.na(genus), genus != "") %>%
  group_by(genus) %>%
  summarise(total_abundance = sum(f_weighted_at_rank, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_abundance)) %>%
  slice_head(n = 10) %>%
  pull(genus)

genus_bar_plot_data <- tax_without_parsed %>%
  mutate(genus_plot = if_else(!is.na(genus) & genus %in% top_genera_bar, genus, "Other")) %>%
  group_by(sample_name, studycode, analysis_group, genus_plot) %>%
  summarise(rel_abundance = sum(f_weighted_at_rank, na.rm = TRUE), .groups = "drop") %>%
  group_by(sample_name, studycode, analysis_group) %>%
  mutate(stack_order = if_else(genus_plot == "Other", -1, rank(rel_abundance, ties.method = "first"))) %>%
  ungroup()

genus_levels <- genus_bar_plot_data %>%
  group_by(genus_plot) %>%
  summarise(total_abundance = sum(rel_abundance), .groups = "drop") %>%
  arrange(desc(total_abundance)) %>%
  pull(genus_plot)

genus_fill <- make_fill_palette(genus_levels)

genus_bar_plot_data <- genus_bar_plot_data %>%
  mutate(
    studycode = factor(studycode, levels = studycode_order),
    genus_plot = factor(genus_plot, levels = genus_levels)
  )

genus_bar_plot <- ggplot(genus_bar_plot_data, aes(x = studycode, y = rel_abundance, fill = genus_plot, order = stack_order)) +
  geom_col(width = 0.75, color = "grey15", linewidth = 0.15) +
  facet_wrap(~ analysis_group, scales = "free_x", labeller = group_labeller) +
  scale_y_continuous(
    labels = percent_format(accuracy = 1),
    limits = c(0, 1),
    expand = expansion(mult = c(0, 0.02))
  ) +
  scale_fill_manual(values = genus_fill, breaks = genus_levels) +
  labs(
    title = "Genus-level composition after removing unclassified rows",
    subtitle = "Top 10 genera by overall relative abundance; 'Other' is stacked on top within each sample",
    x = "Study code",
    y = "Relative abundance (f_weighted_at_rank)",
    fill = "Genus"
  ) +
  theme(
    axis.text.x = element_text(angle = 20, hjust = 1),
    legend.position = "right"
  )

save_fig(genus_bar_plot, "taxonomy_genus_composition.png", width = 10, height = 6)
genus_bar_plot
```

**Take-home note.** The genus-level stacked bar plot is more detailed than the phylum plot and helps show which named genera drive the broad taxonomic shifts.

```{r}
genus_top10_by_sample <- tax_without_parsed %>%
  filter(!is.na(genus), genus != "") %>%
  group_by(studycode, analysis_group, genus) %>%
  summarise(rel_abundance = sum(f_weighted_at_rank, na.rm = TRUE), .groups = "drop") %>%
  group_by(studycode, analysis_group) %>%
  arrange(desc(rel_abundance), .by_group = TRUE) %>%
  slice_head(n = 10) %>%
  ungroup()

genus_top10_by_group <- tax_without_parsed %>%
  filter(!is.na(genus), genus != "") %>%
  group_by(studycode, analysis_group, genus) %>%
  summarise(rel_abundance = sum(f_weighted_at_rank, na.rm = TRUE), .groups = "drop") %>%
  group_by(analysis_group, genus) %>%
  summarise(mean_rel_abundance = mean(rel_abundance), .groups = "drop") %>%
  group_by(analysis_group) %>%
  arrange(desc(mean_rel_abundance), .by_group = TRUE) %>%
  slice_head(n = 10) %>%
  ungroup()

knitr::kable(genus_top10_by_sample, digits = 4, caption = "Top 10 genera by relative abundance within each sample")
knitr::kable(genus_top10_by_group, digits = 4, caption = "Top 10 genera by mean relative abundance within each age group")
```

### Genus heatmap

**Code note**

-   The stacked bar plot above is good for overall composition, but it becomes hard to see smaller differences.
-   This heatmap shows the same data in a matrix view, which makes it easier to spot sample-specific enrichment patterns.

```{r}
top_genera <- tax_without_parsed %>%
  filter(!is.na(genus), genus != "") %>%
  group_by(genus) %>%
  summarise(total_abundance = sum(f_weighted_at_rank, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(total_abundance)) %>%
  slice_head(n = 15) %>%
  pull(genus)

genus_plot_data <- tax_without_parsed %>%
  filter(genus %in% top_genera) %>%
  group_by(sample_name, studycode, analysis_group, genus) %>%
  summarise(rel_abundance = sum(f_weighted_at_rank, na.rm = TRUE), .groups = "drop") %>%
  mutate(
    studycode = factor(studycode, levels = studycode_order),
    genus = factor(genus, levels = rev(top_genera))
  )

genus_fill_max_raw <- max(genus_plot_data$rel_abundance, na.rm = TRUE)
genus_fill_limit <- ((floor((100 * genus_fill_max_raw) / 10) + 1) * 10) / 100
genus_fill_breaks <- seq(0, genus_fill_limit, by = 0.1)

genus_plot <- ggplot(genus_plot_data, aes(x = studycode, y = genus, fill = rel_abundance)) +
  geom_tile(color = "white", linewidth = 0.2) +
  facet_wrap(~ analysis_group, scales = "free_x", labeller = group_labeller) +
  scale_fill_gradient(
    low = "grey95",
    high = "#2C7FB8",
    limits = c(0, genus_fill_limit),
    breaks = genus_fill_breaks,
    labels = percent_format(accuracy = 1)
  ) +
  labs(
    title = "Top genera across the six re-analyzed samples",
    x = "Study code",
    y = "Genus",
    fill = "Relative abundance"
  ) +
  theme(axis.text.x = element_text(angle = 20, hjust = 1))

save_fig(genus_plot, "taxonomy_genus_heatmap.png", width = 10, height = 6.5)
genus_plot
```

**Take-home note.** The heatmap makes it easier to see sample-specific genus enrichment patterns that are harder to spot in the stacked bar plot.

**Descriptive interpretation**

-   The two youngest-child samples show more **Actinomycetota** than the adult samples, driven largely by **Bifidobacterium**.
-   Adult samples are richer in the classic adult-associated **Bacillota + Bacteroidota** balance.
-   The single older-child sample has a phylum profile that looks more adult-like than the two youngest samples, but a single sample is not enough to claim convergence.

## Alpha diversity

**Code note**

-   Alpha diversity asks how diverse each sample is on its own.
-   We use the Shannon index because it captures both richness and evenness and is familiar to many microbiome readers.

```{r}
alpha_diversity <- tax_without_parsed %>%
  group_by(sample_name) %>%
  summarise(
    shannon = vegan::diversity(f_weighted_at_rank, index = "shannon"),
    observed_taxa = n_distinct(match_name),
    .groups = "drop"
  ) %>%
  left_join(metadata, by = "sample_name") %>%
  arrange(analysis_group, studycode)

knitr::kable(alpha_diversity, digits = 3)
```

```{r}
alpha_jitter <- position_jitter(width = 0.06, height = 0, seed = 42)

alpha_plot <- ggplot(alpha_diversity, aes(x = analysis_group, y = shannon, color = analysis_group)) +
  geom_boxplot(width = 0.55, outlier.shape = NA, alpha = 0.2, linewidth = 0.4) +
  geom_point(aes(shape = analysis_group), size = 3, position = alpha_jitter) +
  ggrepel::geom_text_repel(
    aes(label = studycode),
    position = alpha_jitter,
    size = 3,
    show.legend = FALSE,
    seed = 42,
    box.padding = 0.35,
    point.padding = 0.2,
    min.segment.length = 0
  ) +
  scale_color_manual(values = group_palette, labels = group_labels) +
  scale_shape_manual(values = group_shapes, labels = group_labels) +
  scale_x_discrete(labels = group_labels) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  labs(
    #title = "Adults show the highest Shannon diversity in this 6-sample subset",
    x = "Analysis group",
    y = "Shannon diversity",
    color = "Group",
    shape = "Group"
  ) +
  theme(legend.position = "none")

save_fig(alpha_plot, "taxonomy_alpha_diversity.png", width = 5, height = 5)
alpha_plot
```

**Take-home note.** In this subset, adult samples sit at the higher end of Shannon diversity, while the child samples are consistently lower.

No formal groupwise alpha-diversity test is emphasized here. With group sizes of **2**, **1**, and **3**, a p-value would suggest more precision than the data can support.

## Beta diversity / ordination

**Code note**

-   Beta diversity compares samples against one another rather than within a single sample.
-   Here we build a species-by-sample matrix from `f_weighted_at_rank`, compute Bray-Curtis dissimilarity, and then use classical multidimensional scaling to visualize those distances.

```{r}
species_wide <- tax_without_parsed %>%
    select(sample_name, match_name, f_weighted_at_rank) %>%
    distinct() %>%
    pivot_wider(names_from = match_name, values_from = f_weighted_at_rank, values_fill = 0) %>%
    arrange(match(sample_name, sample_order))

species_mat <- as.matrix(as.data.frame(species_wide[, -1]))
rownames(species_mat) <- species_wide$sample_name

bray_dist <- vegdist(species_mat, method = "euclidean")
bray_cmd <- cmdscale(bray_dist, eig = TRUE, k = 2)

beta_diversity <- tibble::tibble(
    sample_name = rownames(bray_cmd$points),
    Axis1 = bray_cmd$points[, 1],
    Axis2 = bray_cmd$points[, 2]
) %>%
    left_join(metadata, by = "sample_name")

positive_eig <- bray_cmd$eig[bray_cmd$eig > 0]
axis1_var <- positive_eig[1] / sum(positive_eig)
axis2_var <- positive_eig[2] / sum(positive_eig)

pairwise_bray <- as.matrix(bray_dist) %>%
    as.data.frame() %>%
    tibble::rownames_to_column("sample_1") %>%
    pivot_longer(-sample_1, names_to = "sample_2", values_to = "bray_curtis") %>%
    filter(sample_1 < sample_2) %>%
    left_join(metadata %>% select(sample_name, group_1 = analysis_group), by = c("sample_1" = "sample_name")) %>%
    left_join(metadata %>% select(sample_name, group_2 = analysis_group), by = c("sample_2" = "sample_name")) %>%
    left_join(metadata %>% select(sample_name, studycode_1 = studycode), by = c("sample_1" = "sample_name")) %>%
    left_join(metadata %>% select(sample_name, studycode_2 = studycode), by = c("sample_2" = "sample_name")) %>%
    select(studycode_1, studycode_2, group_1, group_2, bray_curtis) %>%
    arrange(bray_curtis)

knitr::kable(pairwise_bray, digits = 3)
```

```{r}
beta_plot <- ggplot(beta_diversity, aes(x = Axis1, y = Axis2, color = analysis_group)) +
  geom_point(aes(shape = analysis_group), size = 3) +
  stat_ellipse(aes(group = analysis_group), linewidth = 0.6, linetype = 2, level = 0.8, show.legend = FALSE) +
  ggrepel::geom_text_repel(
    aes(label = studycode),
    size = 3,
    show.legend = FALSE,
    seed = 42,
    box.padding = 0.25,
    point.padding = 0.25,
    min.segment.length = 0
  ) +
  scale_color_manual(values = group_palette, labels = group_labels) +
  scale_shape_manual(values = group_shapes, labels = group_labels) +
  scale_x_continuous(expand = expansion(mult = c(0.05, 0.05))) +
  scale_y_continuous(expand = expansion(mult = c(0.05, 0.05))) +
  labs(
    title = "Bray-Curtis ordination on species-level f_weighted_at_rank",
    x = sprintf("Axis 1 (%.1f%% explained)", 100 * axis1_var),
    y = sprintf("Axis 2 (%.1f%% explained)", 100 * axis2_var),
    color = "Group",
    shape = "Group"
  )

save_fig(beta_plot, "taxonomy_bray_curtis_ordination.png", width = 8.5, height = 5.8)
beta_plot
```

**Take-home note.** The ordination shows that the adult samples group more tightly together, while the child samples are farther away from that adult cluster.

## Taxonomy interpretation

This subset is directionally consistent with the paper in some ways:

-   The **adult samples cluster more closely to one another** than the two youngest-child samples.
-   Adults also show the **highest Shannon diversity** in the subset.
-   The two youngest-child samples are compositionally distinct, especially because of stronger **Actinomycetota / Bifidobacterium** signal and lower Shannon diversity.

However, the subset does **not** fully recapitulate the paper's age gradient:

-   There is only **one** older-child sample.
-   That older-child sample has a more adult-like phylum balance than the younger-child samples, but it still remains Bray-Curtis-distant from the adults.
-   Therefore, we can say the local data are **compatible with** the reported maturation trend, but this subset is too small to claim a clean replication.


::: {.callout-tip}
[Back to workflow setup](vietnamese-healthy-gut-microbiome.qmd) | Continue to [AMR Analysis & Summary](vietnamese-healthy-gut-amr.qmd).
:::