---
title: "Clinical Foundations & AMR Surveillance: AMR Analysis & Summary"
subtitle: "AMR burden, filtering choices, and cautious comparison with Pereira-Dias et al."
author: "Hao Chung The, Minh-Quan Ton-Ngoc"
date: "2026-05-15"
format:
  html:
    toc: true
    toc-depth: 3
    toc-expand: true
    number-sections: true
    code-fold: true
    theme: cosmo
    self-contained: true
    smooth-scroll: true
    embed-resources: true
execute:
  echo: true
  warning: false
  message: false
---

```{=html}
<style>
.manual-note {
  background: #fff8e1;
  border-left: 4px solid #f0ad4e;
  padding: 0.9rem 1rem;
  border-radius: 0.4rem;
  margin: 1rem 0;
}
.analysis-note {
  background: #eef6ff;
  border-left: 4px solid #1f78b4;
  padding: 0.95rem 1.1rem;
  border-radius: 0.4rem;
  margin: 1rem 0;
}
.result-card {
  background: #f6fff8;
  border-left: 4px solid #198754;
  padding: 1rem 1.2rem;
  border-radius: 0.4rem;
  margin: 1rem 0;
}
.debrief-box {
  background: #f7f7fb;
  border: 1px solid #d9dee3;
  border-radius: 0.5rem;
  padding: 1rem 1.2rem;
  margin: 1rem 0;
}
.mcq {
  border: 1px solid #d9dee3;
  border-radius: 0.5rem;
  padding: 1rem;
  margin: 1rem 0 1.25rem 0;
  background: #ffffff;
}
.mcq .question {
  font-weight: 600;
  margin-bottom: 0.75rem;
}
.mcq label {
  display: block;
  margin: 0.35rem 0;
}
.mcq button {
  margin-top: 0.75rem;
  margin-right: 0.5rem;
  padding: 0.35rem 0.75rem;
  border: 1px solid #cfd4da;
  border-radius: 0.35rem;
  background: #f8f9fa;
  color: #212529;
  font-size: 0.95rem;
  line-height: 1.4;
  box-shadow: none;
  cursor: pointer;
}
.mcq button:hover,
.mcq button:focus {
  background: #eef1f4;
  border-color: #adb5bd;
}
.mcq .feedback {
  margin-top: 0.75rem;
  padding: 0.75rem;
  border-radius: 0.35rem;
  display: none;
}
.mcq .feedback.correct {
  display: block;
  background: #e8f5e9;
  color: #1b5e20;
}
.mcq .feedback.incorrect {
  display: block;
  background: #fdecea;
  color: #8a1c1c;
}
.next-step {
  margin-top: 0.75rem;
}
.small-text {
  font-size: 0.95rem;
  color: #495057;
}
summary {
  font-weight: 600;
}
code, pre, pre code {
  white-space: pre-wrap;
  overflow-wrap: anywhere;
  word-break: break-word;
}
p code, li code, td code {
  display: inline-block;
  max-width: 100%;
  vertical-align: bottom;
}
.sidebar nav[role="doc-toc"],
.sidebar nav[role="doc-toc"] a,
#TOC,
#TOC a {
  overflow-wrap: anywhere;
  word-break: break-word;
}
</style>

<script>
const mcqFeedbackGut = {
  mcq_scope: {
    A: "That would overstate what six samples can support. The notebook is deliberately framed as a workflow exercise, not a definitive cohort-level replication.",
    B: "Correct. The notebook is meant to teach a compact end-to-end workflow and support cautious directional comparison, not reproduce the entire paper.",
    C: "The workshop absolutely does use real biological data and real outputs. The point is not to avoid biology, but to scale the exercise to something teachable.",
    D: "The paper already exists, so the workshop does not need to repeat every methodological choice exactly. It needs to be explicit about where the teaching workflow differs."
  },
  mcq_setup: {
    A: "A browser is helpful for ENA and HTML reports, but it is not the core requirement that makes the pipeline runnable.",
    B: "Correct. Nextflow needs Java, and the workflow also needs a way to resolve tool environments such as Conda or Apptainer.",
    C: "R is important later for exploratory analysis, but the pipeline itself starts well before the R section.",
    D: "The workshop databases are already prepared. Learners do not need to rebuild them to get started."
  },
  mcq_csv: {
    A: "This is too minimal for the documented samplesheet format. The current wiki expects the short-read columns and a placeholder long-read column.",
    B: "Correct. The documented CSV uses `sample_id,run_id,group,short_reads_1,short_reads_2,long_reads`, with `run_id` optional and `long_reads` left blank here.",
    C: "These column names are close to what some pipelines use, but they do not match the format described in the current metaflow wiki.",
    D: "A TSV can be useful in other contexts, but the documented input here is a CSV samplesheet."
  },
  mcq_config: {
    A: "Those settings matter, but they are already prepared for the workshop and are not the first things learners should edit.",
    B: "Correct. `input` and `outdir` are the two essential path variables learners are expected to adjust before launching the Vietnamese healthy gut microbiome run.",
    C: "Resource settings can be tuned later if needed, but most learners should not begin there.",
    D: "ARG settings are relevant, but the pipeline will not even know what to process if `input` and `outdir` are wrong."
  },
  mcq_ena: {
    A: "The parent study accession is useful context, but it does not itself give you the paired FASTQ URLs you need for sample download.",
    B: "Correct. `fastq_ftp` is the field that exposes the paired download links, while `sample_title` helps you match the correct workshop samples.",
    C: "Coverage and insert size can be interesting later, but they are not the key fields for collecting the raw FASTQ files in this exercise.",
    D: "The publication title helps you orient yourself, but it is not how you recover the read files."
  },
  mcq_outputs: {
    A: "That would hide an important teaching point. The difference between classified-only and with-unclassified views is part of the interpretation lesson.",
    B: "Correct. The two tables answer slightly different questions: one keeps the uncertainty visible, while the other describes composition conditional on classified signal.",
    C: "The issue is not raw speed. The main reason is interpretability and transparency about unclassified signal.",
    D: "The AMR outputs are useful, but they are not the reason for generating both taxonomy tables."
  },
  mcq_unclassified: {
    A: "That would be too strong. A large unclassified fraction means the named taxa summarize only the classified part of the sample, not the whole metagenome.",
    B: "Correct. Once the unclassified fraction is large, the classified-only table is best read as composition conditional on what the database could label.",
    C: "The problem is not solved by removing rows alone. Removing unclassified rows changes the meaning of the remaining relative abundances.",
    D: "High unclassified signal does not invalidate every downstream step, but it does require more careful wording and comparison."
  }
};

function checkMCQGut(id, correctValue) {
  const box = document.getElementById(id);
  const selected = box.querySelector('input[type="radio"]:checked');
  const feedback = box.querySelector('.feedback');
  if (!selected) {
    feedback.className = 'feedback incorrect';
    feedback.style.display = 'block';
    feedback.innerHTML = 'Please choose one answer before checking.';
    return;
  }
  const isCorrect = selected.value === correctValue;
  feedback.className = isCorrect ? 'feedback correct' : 'feedback incorrect';
  feedback.innerHTML = mcqFeedbackGut[id][selected.value];
  feedback.style.display = 'block';
}

function resetMCQGut(id) {
  const box = document.getElementById(id);
  box.querySelectorAll('input[type="radio"]').forEach(el => el.checked = false);
  const feedback = box.querySelector('.feedback');
  feedback.style.display = 'none';
  feedback.innerHTML = '';
}
</script>
```

# R-based exploratory analysis

The sections below assume that the pipeline outputs above already exist. From this point onward, the notebook shifts from <strong>pipeline execution</strong> to <strong>exploratory interpretation</strong>.

```{r}
library(dplyr)
library(tidyr)
library(readr)
library(ggplot2)
library(ggrepel)
library(phyloseq)
library(vegan)
library(patchwork)
library(scales)
library(data.table)
library(broom)

data_dir <- file.path("..", "materials", "vietnamese-healthy-gut-microbiome")

group_palette <- c(
  "Child 0-23 months" = "#D55E00",
  "Child 24-59 months" = "#0072B2",
  "Adult" = "#009E73"
)

group_shapes <- c(
  "Child 0-23 months" = 16,
  "Child 24-59 months" = 17,
  "Adult" = 15
)

group_labels <- c(
  "Child 0-23 months" = "Child\n(0-23 months)",
  "Child 24-59 months" = "Child\n(24-59 months)",
  "Adult" = "Adult"
)

group_labeller <- labeller(analysis_group = as_labeller(group_labels))

my_color <- c(
  "#4e79a7", "#f28e2b", "#59a14f", "#b6992d",
  "#499894", "#e15759", "#b07aa1", "#9d7660"
)

make_fill_palette <- function(levels_vector, other_label = "Other") {
  core_levels <- setdiff(levels_vector, other_label)
  n_cols <- length(core_levels)
  if (n_cols > length(my_color)) {
    extra_cols <- scales::hue_pal()(n_cols - length(my_color))
    core_cols <- c(my_color, extra_cols)
  } else {
    core_cols <- my_color[seq_len(n_cols)]
  }
  cols <- setNames(core_cols, core_levels)
  if (other_label %in% levels_vector) {
    cols[other_label] <- "grey85"
  }
  cols[levels_vector]
}

paper_amr_class_levels <- c(
  "Aminoglycosides", "β-lactamases", "Fosfomycin", "Fluoroquinolones",
  "Glycopeptides", "MLS", "Chloramphenicol", "Sulfonamides",
  "Tetracyclines", "Trimethoprim", "Other"
)

map_paper_amr_class <- function(drug_class) {
  case_when(
    drug_class %in% c("aminoglycoside antibiotic") ~ "Aminoglycosides",
    drug_class %in% c("cephalosporin", "penicillin beta-lactam", "carbapenem", "monobactam") ~ "β-lactamases",
    drug_class %in% c("phosphonic acid antibiotic") ~ "Fosfomycin",
    drug_class %in% c("fluoroquinolone antibiotic") ~ "Fluoroquinolones",
    drug_class %in% c("glycopeptide antibiotic") ~ "Glycopeptides",
    drug_class %in% c(
      "macrolide antibiotic", "lincosamide antibiotic", "streptogramin antibiotic",
      "streptogramin A antibiotic", "streptogramin B antibiotic"
    ) ~ "MLS",
    drug_class %in% c("phenicol antibiotic") ~ "Chloramphenicol",
    drug_class %in% c("sulfonamide antibiotic") ~ "Sulfonamides",
    drug_class %in% c("tetracycline antibiotic", "glycylcycline") ~ "Tetracyclines",
    drug_class %in% c("diaminopyrimidine antibiotic") ~ "Trimethoprim",
    TRUE ~ "Other"
  )
}

theme_set(
  theme_bw(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      plot.title.position = "plot",
      legend.position = "right"
    )
)

fig_dir <- "figures"
dir.create(fig_dir, showWarnings = FALSE)

save_fig <- function(plot, filename, width = 9, height = 5.5, dpi = 300) {
  ggsave(
    filename = file.path(fig_dir, filename),
    plot = plot,
    width = width,
    height = height,
    dpi = dpi,
    bg = "white"
  )
}

parse_lineage <- function(df) {
  df %>%
    tidyr::separate(
      col = lineage,
      into = c("domain", "phylum", "class", "order", "family", "genus", "species"),
      sep = ";",
      fill = "right",
      extra = "drop",
      remove = FALSE
    )
}

trim_sample_ids <- function(x) trimws(as.character(x))
```

**Code note**

-   This setup chunk loads the R packages used throughout the notebook and defines a few helper objects.
-   The notebook now uses **ggplot2 only** for plotting, to keep figure sizing predictable in the rendered output.
-   For color, the stacked bar plots use a richer custom palette, while group-level plots still combine color and point shape so they remain readable on screen and when printed.
-   The default ggplot2 theme is `theme_bw()`, so any plot built directly with ggplot2 follows the workshop style guide.

# Context from the original paper

Before looking at the 6 local samples, we extract the main analytical context from the published article and the local summary notes.

**Paper-level findings that are reasonable to compare against in a workshop subset**

-   The study analyzed **42 healthy Vietnamese participants** split into **0-23 months**, **2-5 years**, and **adults**.
-   The paper reported a **clear age-related microbiome structure difference**, with children under 2 years showing a distinct gut community structure.
-   **Alpha diversity** was lower in the youngest children than in older children and adults.
-   **Beta diversity** showed broad overlap between older children and adults, while younger children were more distinct and more dispersed.
-   The healthy gut resistome was described as a substantial AMR reservoir, with **tetracycline** and **MLS** resistance genes especially common.
-   Genes highlighted in the paper included **ermB**, **lnuC**, **tetM**, **tetO**, **tetQ**, **tetW**, and **tet-32** as near-ubiquitous signals.
-   The paper reported a higher AMR burden in the youngest children, and noted clinically important genes such as **blaCTX-M** and **mphA/mphD** in younger children.

```{r}
paper_context <- tibble::tibble(
  theme = c(
    "Age-related microbiome structure",
    "Alpha diversity",
    "Beta diversity",
    "Broad AMR burden",
    "AMR classes emphasized",
    "AMR genes emphasized"
  ),
  published_finding = c(
    "Children under 2 years had a distinct gut microbiome structure relative to older children and adults.",
    "Youngest children had lower Shannon diversity; older children were closer to adults.",
    "Bray-Curtis ordination showed young children dispersed away from the overlapping older-child and adult cloud.",
    "Youngest children carried the greatest number of detected AMR genes.",
    "Tetracycline and MLS resistance were nearly ubiquitous.",
    "ermB, lnuC, tet-32, tetM, tetO, tetQ, tetW were common; blaCTX-M and mphA/mphD were highlighted as clinically important signals."
  )
)

knitr::kable(paper_context, col.names = c("Theme", "Finding extracted from paper / summary"))
```

# Analysis-ready files used below

The R analysis below assumes that the workflow and post-processing steps have already produced an analysis folder containing:

-   Metadata for the 6 re-analyzed samples.
-   Taxonomy tables with and without unclassified rows.
-   A `vietnamese_gut.phyloseq` folder that contains `otu_table.csv`, `sample_data.csv`, and `tax_table.csv`.
-   A `vietnamese_gut_amr` folder with one `rgi-bwt` gene mapping table per sample.
-   The original paper PDF and a local summary file used only for context.

```{r}
base_dir <- "."

paths <- tibble::tibble(
  object = c(
    "metadata",
    "taxonomy_with_unclassified",
    "taxonomy_without_unclassified",
    "phyloseq_export_dir",
    "amr_dir",
    "local_summary_md"
  ),
  file_name = c(
    "vietnamese_gut_sample_information.csv",
    "vietnamese_gut_with_unclassified.csv",
    "vietnamese_gut_without_unclassified.csv",
    "vietnamese_gut.phyloseq",
    "vietnamese_gut_amr",
    "vietnamese_gut_summary.md"
  )
) %>%
  mutate(
  path = file.path(data_dir, file_name),
  exists = file.exists(path) | dir.exists(path)
)

knitr::kable(paths)
```

# Metadata inspection

## Load data

**Code note**

-   This chunk reads the local sample metadata and creates a simple `analysis_group` variable that we can use consistently in plots and summaries.
-   The grouping is intentionally simple because the metadata for this workshop subset are limited.

```{r}
metadata <- read_csv(file.path(data_dir, "vietnamese_gut_sample_information.csv"), show_col_types = FALSE) %>%
  mutate(
    sample_name = trim_sample_ids(sample_name),
    studycode = trim_sample_ids(studycode),
    status = recode(status, "Children" = "Child", .default = status),
    analysis_group = case_when(
      status == "Adult" ~ "Adult",
      age_month < 24 ~ "Child 0-23 months",
      TRUE ~ "Child 24-59 months"
    ),
    analysis_group = factor(
      analysis_group,
      levels = c("Child 0-23 months", "Child 24-59 months", "Adult")
    )
  ) %>%
  arrange(analysis_group, studycode)

sample_order <- metadata %>%
  arrange(analysis_group, studycode) %>%
  pull(sample_name)

studycode_order <- metadata %>%
  arrange(analysis_group, studycode) %>%
  pull(studycode)

metadata
```

**Metadata notes**

-   `status` is the most reliable top-level grouping variable in the local file.
-   For children, `age_month` is usable as age in months.
-   For adults, the `age_month` column appears to contain values such as 25, 32, and 59, which are much more plausible as adult ages in **years** than months.
-   For that reason, this notebook uses `analysis_group` derived from `status` and child age thresholds, and uses the adult age values only as descriptive metadata.

```{r}
metadata_summary <- metadata %>%
  count(status, analysis_group, name = "n_samples")

knitr::kable(metadata_summary)
```

## Harmonize sample IDs across inputs

**Code note**

-   This chunk checks whether the same sample IDs are present in metadata, taxonomy tables, the exported phyloseq tables, and the AMR result files.
-   Harmonizing IDs early helps avoid silent joins to the wrong sample later in the notebook.
-   We keep `sample_name` for joins, but use `studycode` later for plotting because it is more compact on figure axes.

```{r}
tax_with <- read_csv(file.path(data_dir, "vietnamese_gut_with_unclassified.csv"), show_col_types = FALSE) %>%
  mutate(sample_name = trim_sample_ids(sample_name))

tax_without <- read_csv(file.path(data_dir, "vietnamese_gut_without_unclassified.csv"), show_col_types = FALSE) %>%
  mutate(sample_name = trim_sample_ids(sample_name))

phy_sample_data <- read_csv(file.path(data_dir, "vietnamese_gut.phyloseq", "sample_data.csv"), show_col_types = FALSE) %>%
  mutate(sample_name = trim_sample_ids(sample_name))

amr_files <- list.files(file.path(data_dir, "vietnamese_gut_amr"), pattern = "gene_mapping_data\\.txt$", full.names = TRUE)
amr_file_samples <- sub("\\.gene_mapping_data\\.txt$", "", basename(amr_files))

sample_harmonization <- tibble::tibble(
  source = c("metadata", "taxonomy_with", "taxonomy_without", "phyloseq_export", "amr_files"),
  n_unique_samples = c(
    n_distinct(metadata$sample_name),
    n_distinct(tax_with$sample_name),
    n_distinct(tax_without$sample_name),
    n_distinct(phy_sample_data$sample_name),
    length(unique(amr_file_samples))
  ),
  sample_ids = c(
    paste(sort(unique(metadata$sample_name)), collapse = ", "),
    paste(sort(unique(tax_with$sample_name)), collapse = ", "),
    paste(sort(unique(tax_without$sample_name)), collapse = ", "),
    paste(sort(unique(phy_sample_data$sample_name)), collapse = ", "),
    paste(sort(unique(amr_file_samples)), collapse = ", ")
  )
)

knitr::kable(sample_harmonization)
```

All four sources contain the same six sample IDs, so harmonization is straightforward.

## Inspect the exported phyloseq components

```{r}
phy_otu_export <- read_csv(file.path(data_dir, "vietnamese_gut.phyloseq", "otu_table.csv"), show_col_types = FALSE)
phy_tax_export <- read_csv(file.path(data_dir, "vietnamese_gut.phyloseq", "tax_table.csv"), show_col_types = FALSE)

phyloseq_export_status <- tibble::tibble(
  file = c("sample_data.csv", "otu_table.csv", "tax_table.csv"),
  rows = c(nrow(phy_sample_data), nrow(phy_otu_export), nrow(phy_tax_export)),
  columns = c(ncol(phy_sample_data), ncol(phy_otu_export), ncol(phy_tax_export))
)

knitr::kable(phyloseq_export_status)
```

The `vietnamese_gut.phyloseq` path is not a serialized `.rds` phyloseq object. It is an export directory containing the component tables needed to reconstruct a phyloseq object.

## Inspect the AMR folder layout

```{r}
amr_file_inventory <- tibble::tibble(
  file = basename(amr_files),
  sample_name = sub("\\.gene_mapping_data\\.txt$", "", basename(amr_files)),
  size_bytes = file.info(amr_files)$size
)

knitr::kable(amr_file_inventory)
```

```{r}
amr_example <- fread(amr_files[1], sep = "\t", data.table = FALSE)

knitr::kable(head(amr_example, 8))
```

The AMR files are one table per sample and include:

-   ARO term
-   reference model metadata
-   mapped read counts
-   average coverage
-   AMR gene family
-   drug class
-   resistance mechanism

This is sufficient for descriptive gene-level and drug-class summaries.

# AMR analysis

## Inspect rgi-bwt outputs and harmonize sample IDs

**Code note**

-   This chunk reads all per-sample `rgi-bwt` result tables into one long table.
-   It also converts the key AMR columns to numeric form and extracts a usable gene length from `Reference Length`.
-   We then build **two normalized AMR datasets**:
    1.  a version without a coverage filter
    2.  a stricter version retaining only hits with `Average Percent Coverage ≥ 90`

```{r}
amr_raw <- lapply(amr_files, function(path) {
  read_tsv(path, show_col_types = FALSE) %>%
    mutate(
      sample_name = sub("\\.gene_mapping_data\\.txt$", "", basename(path)),
      sample_name = trim_sample_ids(sample_name)
    )
}) %>%
  bind_rows() %>%
  left_join(metadata, by = "sample_name") %>%
  mutate(
    `Completely Mapped Reads` = as.numeric(`Completely Mapped Reads`),
    `All Mapped Reads` = as.numeric(`All Mapped Reads`),
    `Average Percent Coverage` = as.numeric(`Average Percent Coverage`),
    `Average Length Coverage (bp)` = as.numeric(`Average Length Coverage (bp)`),
    reference_length_raw = as.character(`Reference Length`),
    reference_length_bp = readr::parse_number(sub(";.*$", "", reference_length_raw))
  )

amr_base <- amr_raw %>%
  filter(
    !is.na(`Completely Mapped Reads`),
    !is.na(reference_length_bp),
    reference_length_bp > 0,
    `Completely Mapped Reads` > 0
  ) %>%
  mutate(
    rpk = `Completely Mapped Reads` / (reference_length_bp / 1000)
  )

amr_all <- amr_base %>%
  group_by(sample_name) %>%
  mutate(
    retained_amr_reads = sum(`Completely Mapped Reads`, na.rm = TRUE),
    retained_amr_reads_million = retained_amr_reads / 1e6,
    rpkm = rpk / retained_amr_reads_million
  ) %>%
  ungroup() %>%
  mutate(amr_case = "No coverage filter")

amr_filtered <- amr_base %>%
  filter(!is.na(`Average Percent Coverage`), `Average Percent Coverage` >= 90) %>%
  group_by(sample_name) %>%
  mutate(
    retained_amr_reads = sum(`Completely Mapped Reads`, na.rm = TRUE),
    retained_amr_reads_million = retained_amr_reads / 1e6,
    rpkm = rpk / retained_amr_reads_million
  ) %>%
  ungroup() %>%
  mutate(amr_case = "Average Percent Coverage ≥ 90")

amr_cases <- bind_rows(amr_all, amr_filtered) %>%
  mutate(
    amr_case = factor(amr_case, levels = c("No coverage filter", "Average Percent Coverage ≥ 90"))
  )

amr_overview <- bind_rows(
  amr_all %>%
    summarise(
      amr_case = "No coverage filter",
      n_rows = n(),
      n_samples = n_distinct(sample_name),
      n_aro_terms = n_distinct(`ARO Term`)
    ),
  amr_filtered %>%
    summarise(
      amr_case = "Average Percent Coverage ≥ 90",
      n_rows = n(),
      n_samples = n_distinct(sample_name),
      n_aro_terms = n_distinct(`ARO Term`)
    )
) %>%
  ungroup()

knitr::kable(amr_overview)
```

Before any downstream AMR summary, this notebook applies the same normalization logic in two different ways:

-   compute **RPK** by normalizing **`Completely Mapped Reads`** by **`Reference Length`**
-   compute **RPKM** by scaling RPK within each sample by the retained per-sample AMR read total
-   compare results **with** and **without** the additional filter **`Average Percent Coverage ≥ 90`**

In other words, the number of reads mapped to each gene (**RGI `Completely Mapped Reads`**) is normalized by the gene length (**RGI `Reference Length`**) to create a **Reads per kilobase of AMR gene (RPK)** metric. That RPK value is then converted to **RPKM** using the sample-specific total retained AMR reads.

$$
\mathrm{RPK}_i = \frac{C_i}{L_i / 1000}
$$

$$
\mathrm{RPKM}_i = \frac{\mathrm{RPK}_i}{\left(\sum_{j=1}^{n} C_j\right) / 10^6}
$$

-   $C_i$: `Completely Mapped Reads` for AMR gene $i$
-   $L_i$: `Reference Length` for AMR gene $i$, in base pairs
-   $\sum_{j=1}^{n} C_j$: total `Completely Mapped Reads` across all retained AMR genes in the same sample
-   $n$: number of retained AMR genes in that sample

This makes it possible for workshop participants to compare the effect of the coverage threshold directly, instead of seeing only one filtered view.

## Per-sample gene burden

**Code note**

-   This section summarizes both AMR cases at the sample level.
-   The main idea is simple: count how many AMR hits remain per sample in each case, then compare their total normalized signal using `RPKM`.

```{r}
amr_sample_summary <- amr_cases %>%
  group_by(amr_case, sample_name) %>%
  summarise(
    n_retained_aro_terms = n_distinct(`ARO Term`),
    total_completely_mapped_reads = sum(`Completely Mapped Reads`, na.rm = TRUE),
    total_rpk = sum(rpk, na.rm = TRUE),
    total_rpkm = sum(rpkm, na.rm = TRUE),
    median_rpkm = median(rpkm, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  left_join(metadata, by = "sample_name") %>%
  arrange(amr_case, analysis_group, studycode)

knitr::kable(amr_sample_summary, digits = 2)
```

```{r}
amr_jitter <- position_jitter(width = 0.05, height = 0, seed = 42)

amr_gene_count_plot <- ggplot(amr_sample_summary, aes(x = analysis_group, y = n_retained_aro_terms, color = analysis_group)) +
  stat_summary(fun = median, geom = "crossbar", width = 0.5, color = "grey30") +
  geom_point(aes(shape = analysis_group), size = 3, position = amr_jitter) +
  ggrepel::geom_text_repel(
    aes(label = studycode),
    position = amr_jitter,
    size = 3,
    show.legend = FALSE,
    seed = 42,
    box.padding = 0.25,
    point.padding = 0.2,
    min.segment.length = 0
  ) +
  scale_color_manual(values = group_palette, labels = group_labels) +
  scale_shape_manual(values = group_shapes, labels = group_labels) +
  facet_wrap(~ amr_case, nrow = 1) +
  scale_x_discrete(labels = group_labels) +
  scale_y_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "AMR hit counts by sample under two filtering strategies",
    x = "Analysis group",
    y = "Number of retained ARO terms",
    shape = "Group"
  ) +
  theme(legend.position = "none")

amr_total_rpkm_plot <- ggplot(amr_sample_summary, aes(x = analysis_group, y = total_rpkm, color = analysis_group)) +
  stat_summary(fun = median, geom = "crossbar", width = 0.5, color = "grey30") +
  geom_point(aes(shape = analysis_group), size = 3, position = amr_jitter) +
  ggrepel::geom_text_repel(
    aes(label = studycode),
    position = amr_jitter,
    size = 3,
    show.legend = FALSE,
    seed = 42,
    box.padding = 0.25,
    point.padding = 0.2,
    min.segment.length = 0
  ) +
  scale_color_manual(values = group_palette, labels = group_labels) +
  scale_shape_manual(values = group_shapes, labels = group_labels) +
  facet_wrap(~ amr_case, nrow = 1) +
  scale_x_discrete(labels = group_labels) +
  scale_y_continuous(labels = comma_format(), limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Total retained AMR signal after RPKM normalization",
    x = "Analysis group",
    y = "Summed RPKM",
    shape = "Group"
  ) +
  theme(legend.position = "none")

amr_burden_plot <- amr_gene_count_plot / amr_total_rpkm_plot
save_fig(amr_burden_plot, "amr_burden_summary.png", width = 9, height = 9.5)
amr_burden_plot
```

**Take-home note.** Comparing the top and bottom panels makes it easier to see that the strict coverage filter reduces both the number of retained hits and the apparent AMR burden.

**Interpretation**

-   Without the coverage filter, the AMR summary is broader and tends to preserve more of the signal that directionally resembles the paper.
-   After applying **`Average Percent Coverage ≥ 90`**, the retained AMR set becomes much smaller and more conservative.
-   The child samples still tend to show higher total AMR signal on average, but the difference becomes weaker after strict filtering, and the age pattern remains exploratory rather than definitive.

## Gene-level summaries

```{r}
amr_gene_summary <- amr_cases %>%
  group_by(amr_case, `ARO Term`) %>%
  summarise(
    n_samples = n_distinct(sample_name),
    total_rpk = sum(rpk, na.rm = TRUE),
    total_rpkm = sum(rpkm, na.rm = TRUE),
    max_rpkm_in_one_sample = max(rpkm, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(amr_case, desc(n_samples), desc(total_rpkm))

knitr::kable(amr_gene_summary %>% group_by(amr_case) %>% slice_head(n = 20), digits = 2)
```

The top gene list changes substantially after strict filtering. Several genes that are prominent without a coverage cutoff do not remain after the **Coverage ≥ 90** rule, so the workshop interpretation below compares both cases directly.

```{r}
paper_theme_genes <- c("ErmB", "lnuC", "tet(32)", "tet(M)", "tet(O)", "tet(Q)", "tet(W)", "mphA")
detected_ctxm_variants <- sort(unique(amr_cases$`ARO Term`[grepl("CTX-M", amr_cases$`ARO Term`, ignore.case = TRUE)]))
focus_gene_set <- unique(c(paper_theme_genes, detected_ctxm_variants))

amr_focus_gene_table <- amr_cases %>%
  filter(`ARO Term` %in% focus_gene_set) %>%
  group_by(amr_case, sample_name, `ARO Term`) %>%
  summarise(
    rpk = sum(rpk, na.rm = TRUE),
    rpkm = sum(rpkm, na.rm = TRUE),
    max_coverage = max(`Average Percent Coverage`, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  complete(
    amr_case = levels(amr_cases$amr_case),
    sample_name = sample_order,
    `ARO Term` = focus_gene_set,
    fill = list(rpk = 0, rpkm = 0, max_coverage = NA_real_)
  )

focus_gene_order <- amr_focus_gene_table %>%
  group_by(amr_case, `ARO Term`) %>%
  summarise(total_rpkm = sum(rpkm, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = amr_case, values_from = total_rpkm, values_fill = 0) %>%
  mutate(delta_rpkm = `No coverage filter` - `Average Percent Coverage ≥ 90`) %>%
  arrange(desc(abs(delta_rpkm)), desc(`No coverage filter`), `ARO Term`) %>%
  pull(`ARO Term`)

amr_focus_gene_table <- amr_focus_gene_table %>%
  left_join(metadata %>% select(sample_name, studycode, analysis_group), by = "sample_name") %>%
  mutate(
    studycode = factor(studycode, levels = studycode_order),
    amr_case = factor(amr_case, levels = levels(amr_cases$amr_case)),
    `ARO Term` = factor(`ARO Term`, levels = rev(focus_gene_order))
  )

gene_fill_max <- max(log10(amr_focus_gene_table$rpkm + 1), na.rm = TRUE)
gene_fill_breaks <- pretty(c(0, gene_fill_max), n = 5)

amr_focus_gene_plot <- ggplot(amr_focus_gene_table, aes(x = studycode, y = `ARO Term`, fill = log10(rpkm + 1))) +
  geom_tile(color = "grey70", linewidth = 0.25) +
  facet_grid(
    amr_case ~ analysis_group,
    scales = "free_x",
    space = "free_x",
    labeller = labeller(analysis_group = as_labeller(group_labels))
  ) +
  scale_fill_gradientn(
    colours = c("grey99", "#FEE08B", "#F46D43", "#A50026"),
    limits = c(0, gene_fill_max),
    breaks = gene_fill_breaks
  ) +
  labs(
    title = "Paper-themed AMR genes under both filtering strategies",
    subtitle = "Rows share one color scale; pale bottom-row tiles indicate signals removed by the Coverage ≥ 90 filter",
    x = "Study code",
    y = "ARO term",
    fill = "log10(RPKM + 1)"
  ) +
  theme(
    #axis.text.x = element_text(angle = 0, hjust = 0),
    axis.text.y = element_text(size = 8),
    strip.background = element_rect(fill = "grey92", color = "black"),
    strip.text = element_text(face = "bold"),
    panel.spacing = grid::unit(0.5, "lines")
  )

save_fig(amr_focus_gene_plot, "amr_paper_theme_gene_heatmap.png", width = 10, height = 8)
amr_focus_gene_plot
```

**Take-home note.** Because both rows use the same fill scale and the genes are ordered by how much signal is lost after filtering, the coverage effect is much easier to read at a glance.

```{r}
amr_focus_gene_summary <- amr_cases %>%
  filter(`ARO Term` %in% focus_gene_set) %>%
  group_by(amr_case, `ARO Term`) %>%
  summarise(
    n_samples = n_distinct(sample_name),
    total_rpk = sum(rpk, na.rm = TRUE),
    total_rpkm = sum(rpkm, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(amr_case, desc(n_samples), desc(total_rpkm))

knitr::kable(amr_focus_gene_summary, digits = 2)
```

## Drug-class summaries

```{r}
present_paper_classes <- amr_cases %>%
  select(`Drug Class`) %>%
  separate_rows(`Drug Class`, sep = ";\\s*") %>%
  mutate(paper_class = map_paper_amr_class(`Drug Class`)) %>%
  distinct(paper_class) %>%
  filter(!is.na(paper_class)) %>%
  mutate(paper_class = factor(paper_class, levels = paper_amr_class_levels)) %>%
  arrange(paper_class) %>%
  pull(paper_class)

amr_class_summary <- amr_cases %>%
  select(amr_case, sample_name, studycode, analysis_group, `ARO Term`, `Drug Class`, rpk, rpkm) %>%
  separate_rows(`Drug Class`, sep = ";\\s*") %>%
  mutate(paper_class = map_paper_amr_class(`Drug Class`)) %>%
  group_by(amr_case, sample_name, studycode, analysis_group, paper_class) %>%
  summarise(
    total_rpk = sum(rpk, na.rm = TRUE),
    total_rpkm = sum(rpkm, na.rm = TRUE),
    n_genes = n_distinct(`ARO Term`),
    .groups = "drop"
  ) %>%
  complete(
    amr_case = levels(amr_cases$amr_case),
    sample_name = sample_order,
    paper_class = present_paper_classes,
    fill = list(total_rpk = 0, total_rpkm = 0, n_genes = 0)
  ) %>%
  left_join(metadata %>% select(sample_name, studycode, analysis_group), by = "sample_name") %>%
  mutate(
    studycode = coalesce(studycode.x, studycode.y),
    analysis_group = coalesce(analysis_group.x, analysis_group.y)
  ) %>%
  select(-studycode.x, -studycode.y, -analysis_group.x, -analysis_group.y) %>%
  mutate(amr_case = factor(amr_case, levels = levels(amr_cases$amr_case)))

amr_class_order <- amr_class_summary %>%
  group_by(amr_case, paper_class) %>%
  summarise(total_rpkm = sum(total_rpkm, na.rm = TRUE), .groups = "drop") %>%
  pivot_wider(names_from = amr_case, values_from = total_rpkm, values_fill = 0) %>%
  mutate(delta_rpkm = `No coverage filter` - `Average Percent Coverage ≥ 90`) %>%
  arrange(desc(abs(delta_rpkm)), desc(`No coverage filter`), paper_class) %>%
  pull(paper_class)

amr_class_summary <- amr_class_summary %>%
  mutate(
    studycode = factor(studycode, levels = studycode_order),
    paper_class = factor(paper_class, levels = rev(amr_class_order))
  )

class_fill_max <- max(log10(amr_class_summary$total_rpkm + 1), na.rm = TRUE)
class_fill_breaks <- pretty(c(0, class_fill_max), n = 5)

amr_class_plot <- ggplot(amr_class_summary, aes(x = studycode, y = paper_class, fill = log10(total_rpkm + 1))) +
  geom_tile(color = "grey70", linewidth = 0.25) +
  facet_grid(
    amr_case ~ analysis_group,
    scales = "free_x",
    space = "free_x",
    labeller = labeller(analysis_group = as_labeller(group_labels))
  ) +
  scale_fill_gradientn(
    colours = c("grey99", "#D1E5F0", "#67A9CF", "#2166AC"),
    limits = c(0, class_fill_max),
    breaks = class_fill_breaks
  ) +
  labs(
    title = "Paper-style AMR classes under both filtering strategies",
    subtitle = "Rows share one color scale, so pale bottom-row tiles highlight class-level signal lost after strict filtering",
    x = "Study code",
    y = "Paper-style AMR class",
    fill = "log10(RPKM + 1)"
  ) +
  theme(
    #axis.text.x = element_text(angle = 45, hjust = 1),
    strip.background = element_rect(fill = "grey92", color = "black"),
    strip.text = element_text(face = "bold"),
    panel.spacing = grid::unit(0.5, "lines")
  )

save_fig(amr_class_plot, "amr_drug_class_heatmap.png", width = 10, height = 7.5)
amr_class_plot
```

**Take-home note.** Showing every class-sample combination, including zeros, makes the class-level difference between the no-filter and coverage-filter views much more explicit.

Only the paper-style AMR classes that are actually present in this 6-sample re-analysis are shown here; classes absent from the current data are omitted to avoid implying that they were observed.

```{r}
amr_class_overall <- amr_class_summary %>%
  group_by(amr_case, paper_class) %>%
  summarise(
    n_samples = n_distinct(sample_name),
    total_rpk = sum(total_rpk),
    total_rpkm = sum(total_rpkm),
    .groups = "drop"
  ) %>%
  arrange(amr_case, desc(total_rpkm))

knitr::kable(amr_class_overall, digits = 2)
```

## Clinically highlighted signals

```{r}
critical_hits <- amr_cases %>%
  filter(grepl("CTX-M|mphA", `ARO Term`, ignore.case = TRUE)) %>%
  select(amr_case, studycode, analysis_group, `ARO Term`, `Completely Mapped Reads`, `Average Percent Coverage`, rpk, rpkm, `Drug Class`) %>%
  arrange(amr_case, `ARO Term`, desc(`Completely Mapped Reads`))

knitr::kable(critical_hits, digits = 2)
```

**AMR interpretation**

-   Without the coverage filter, the AMR profile is broader and closer to the kind of descriptive signal emphasized in the paper.
-   Once **`Average Percent Coverage ≥ 90`** is imposed, the profile becomes much more conservative and several previously visible genes disappear.
-   In the paper-style class grouping, **MLS** and **β-lactamase-related** signals remain prominent, but the balance between classes changes substantially between the two AMR cases.
-   Among the highlighted genes, some signals remain only in the no-filter case, while the strict-filter case retains a much smaller subset, including only limited **CTX-M** signal and no retained `mphA`.
-   This side-by-side comparison is useful for teaching how sensitive resistome interpretation can be to filtering choices.


```{r include=FALSE}
tax_without_parsed <- parse_lineage(tax_without) %>%
  left_join(metadata, by = "sample_name")

alpha_diversity <- tax_without_parsed %>%
  group_by(sample_name) %>%
  summarise(
    shannon = vegan::diversity(f_weighted_at_rank, index = "shannon"),
    observed_taxa = n_distinct(match_name),
    .groups = "drop"
  ) %>%
  left_join(metadata, by = "sample_name") %>%
  arrange(analysis_group, studycode)
```

# Integrated exploratory notes

The next plot is deliberately exploratory. It asks whether samples with lower taxonomic diversity also tend to show a larger AMR signal under the two AMR filtering strategies.

```{r}
integrated_summary <- alpha_diversity %>%
  select(sample_name, studycode, analysis_group, shannon) %>%
  left_join(amr_sample_summary %>% select(amr_case, sample_name, total_rpkm, n_retained_aro_terms), by = "sample_name")

integrated_plot <- ggplot(integrated_summary, aes(x = shannon, y = total_rpkm, color = analysis_group, label = studycode)) +
  geom_point(aes(shape = analysis_group), size = 3) +
  ggrepel::geom_text_repel(
    size = 3,
    show.legend = FALSE,
    seed = 42,
    box.padding = 0.25,
    point.padding = 0.25,
    min.segment.length = 0
  ) +
  scale_color_manual(values = group_palette, labels = group_labels) +
  scale_shape_manual(values = group_shapes, labels = group_labels) +
  facet_wrap(~ amr_case, nrow = 1) +
  scale_x_continuous(limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  scale_y_continuous(labels = comma_format(), limits = c(0, NA), expand = expansion(mult = c(0, 0.05))) +
  labs(
    title = "Exploratory view: AMR RPKM versus Shannon diversity",
    x = "Shannon diversity",
    y = "Summed AMR RPKM",
    color = "Group",
    shape = "Group"
  )

save_fig(integrated_plot, "integrated_amr_vs_alpha.png", width = 9, height = 5)
integrated_plot
```

This plot suggests that the lower-diversity child samples also have relatively high retained AMR signal in both AMR summaries, but the strength of that pattern depends on the filter used. With only six samples, this should be treated as a **visual prompt for a larger follow-up analysis**, not as an inferred biological relationship.

**Take-home note.** The relationship between diversity and AMR signal looks broadly similar in both AMR cases, but the apparent strength of the pattern depends on the filtering rule.

# Comparison with the original paper

```{r}
comparison_table <- tibble::tibble(
  theme = c(
    "Age-related microbiome structure",
    "Alpha diversity pattern",
    "Beta diversity pattern",
    "Broad AMR burden",
    "Key AMR classes",
    "Clinically highlighted genes"
  ),
  paper = c(
    "Young children under 2 years were structurally distinct from older children and adults.",
    "Youngest children had lower diversity; older children approached adult values.",
    "Older children and adults overlapped more than younger children.",
    "Youngest children had the highest median AMR gene burden.",
    "Tetracycline and MLS signals were widespread.",
    "blaCTX-M and mphA/mphD were emphasized in younger children."
  ),
  subset_reanalysis = c(
    "The two youngest-child samples are distinct from the adult cluster; the single older-child sample is intermediate in phylum composition but not fully adult-like.",
    "Adults have the highest Shannon values in this subset; all three child samples are lower.",
    "Adults cluster closer to one another; child samples are more separated from the adults than adults are from each other.",
    "Without the coverage filter, children still tend to show higher AMR signal on average; with the ≥ 90 filter that separation becomes weaker and the retained hit counts do not follow a clean age gradient.",
    "Using paper-style class grouping, MLS and β-lactamase-related signals remain visible in both analyses, but the class balance changes substantially after strict filtering.",
    "CTX-M and mphA-like signals are more visible without the coverage filter; only limited CTX-M signal remains after strict filtering, and mphA is not retained."
  ),
  take_home = c(
    "Broad directional concordance, but only partial support because n is very small.",
    "Directionally consistent with the paper.",
    "Directionally consistent for adult clustering; weaker support for an older-child/adult overlap claim.",
    "Suggestive but not a clean reproduction.",
    "Concordance is stronger in the no-filter view and weaker in the strict-filter view.",
    "Agreement for these clinically highlighted genes is highly filter-dependent."
  )
)

knitr::kable(comparison_table)
```

## Quick summary in English

**Findings that remain broadly similar to the original paper**

-   On the **taxonomy** side, the adult samples in this 6-sample subset still show **higher alpha diversity** and **tighter clustering** than the younger-child samples.
-   The two **children under 24 months** still look compositionally distinct from the adult cluster, which remains directionally consistent with the paper's main microbiome maturation result.
-   The overall structure of the subset is therefore still compatible with the broad idea that younger gut microbiomes are less mature and more distinct than adult microbiomes.

**Findings that differ, weaken, or only partially agree once AMR filtering is tightened**

-   We still cannot fully evaluate the paper's “older children become more adult-like” pattern because this subset contains only **one child aged 24-59 months**.
-   On the **beta-diversity** side, adults cluster more tightly, but the single older-child sample does not clearly overlap with the adult cluster the way the full paper suggested.
-   In the AMR analysis, the no-filter view and the strict-filter view do not tell exactly the same story.
-   Under the stricter workflow, many canonical paper-highlighted genes drop out, while a smaller set of higher-coverage hits remains.
-   This means that AMR concordance with the paper is **stronger descriptively without the coverage filter** and **weaker but more conservative with the ≥ 90 filter**.

**Short take-home message**

-   This 6-sample re-analysis still supports the paper at the **broad microbiome-structure level**.
-   On the AMR side, agreement becomes **stronger in the no-filter descriptive view** and **weaker but more conservative** once the analysis is restricted to **high-coverage hits and RPKM-based summaries**.
-   The comparison should therefore be read as a **workshop-oriented directional cross-check**, not as a full replication of the original paper.

# Limitations

This notebook should be read with several major limitations in mind:

-   It is based on **only six re-analyzed samples**.
-   The original paper analyzed **42 participants** and had more balanced age strata.
-   The local metadata are sparse, and even the `age_month` field is ambiguous for adults.
-   Taxonomic ecology here uses **`f_weighted_at_rank`**, not the paper's original Kraken2/Bracken abundance outputs.
-   The unclassified fraction is large in every sample, so the "without unclassified" analyses describe only the classified component.
-   The AMR workflow is **not directly equivalent** to the paper's SRST2 + ARGANNOT pipeline.
-   The notebook compares two AMR views: one without a coverage filter and one with **`Average Percent Coverage ≥ 90`**. The stricter view is conservative and may remove biologically real but lower-coverage hits.
-   The AMR normalization uses **`Completely Mapped Reads`**, **`Reference Length`**, and a sample-internal RPKM scaling based on retained AMR reads available in the local RGI tables.
-   Because the local AMR tables do not include total metagenomic read counts, this RPKM calculation is an available-data approximation rather than a perfect reconstruction of the paper's SRST2-based RPKM framework.
-   With group sizes of **2**, **1**, and **3**, formal inference is not the main point of the exercise.

# Suggested next analyses

If this workshop analysis were expanded into a more complete re-analysis, the next priorities would be:

1.  Re-analyze a much larger subset, ideally recovering age groups closer to the original paper design.
2.  Standardize metadata carefully, especially adult age, diet, household structure, and recent antimicrobial exposure.
3.  Re-run taxonomic profiling with a workflow closer to the original study if strict replication is the goal.
4.  Add a more principled AMR comparison layer, for example family-level harmonization between CARD and ARGANNOT outputs.
5.  Investigate whether the strongest CTX-M detections remain after stricter filtering on coverage and breadth.
6.  Explore organism-to-gene linkage methods if the training objective extends beyond simple resistome description.

# Conclusion

This 6-sample workshop re-analysis captures several of the paper's broad ideas. The taxonomy results remain directionally compatible with a more adult-like and more diverse microbiome in adults.

At the same time, the subset does not cleanly reproduce every published AMR pattern, and the answer depends strongly on which AMR view is used. The **no-filter normalized view** preserves more of the broad descriptive signal that resembles the paper, whereas the **high-coverage view** is more conservative and retains a smaller subset of genes and class-level patterns. Because the sample count is tiny and the AMR detection workflow differs, these differences should be interpreted as a reminder about **workflow dependence, filtering sensitivity, and sampling uncertainty**, not as evidence that the original paper was wrong.

The main workshop take-home is that this small re-analysis is useful for learning how to structure a metagenomic notebook, inspect taxonomy with and without unclassified signal, summarize AMR at gene and drug-class levels, and compare cautiously back to a published study. It is **not** sufficient for definitive biological conclusions.

::: {.callout-tip}
Continue to [Actionable Diagnostics & Clinical Reasoning](clinical-reasoning.qmd).
:::
